# Short-term rentals in Toronto: Housing impacts

<style type="text/css">
  body{
  font-family: Futura, Helvetica, Arial;
}
</style>

<br>

```{r setup, include = FALSE, echo = FALSE}

###### for max to be able to knit
if(memory.limit()< 20000) {memory.limit(size = 48000)}
####

library(here)
source(here("R", "01_startup.R"))

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(include = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# Function for checking file modifications
mtime <- function(files) lapply(Sys.glob(files), function(x) file.info(x)$mtime)

# Set this to TRUE to output PDF figures
build_figures <- FALSE

```

```{r new_objects, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

qload(here("output", "str_processed.qsm"), nthreads = availableCores())
qload(here("output", "geometry.qsm"), nthreads = availableCores())

FREH <- 
  daily %>% 
  filter(date >= "2016-01-01") %>% 
  group_by(date) %>% 
  summarize(across(c(FREH, FREH_3), sum)) %>%
  filter(substr(date, 9, 10) == "01")

FREH_WD <- 
  daily %>% 
  filter(date %in% as.Date(c("2019-12-31", "2020-12-31"))) %>% 
  group_by(ward, date) %>% 
  summarize(FREH = sum(FREH_3)) %>% 
  mutate(date = if_else(date >= start_2020, 2020, 2019))

FREH_DA <- 
  daily %>% 
  filter(date == "2020-12-01") %>% 
  left_join(select(st_drop_geometry(property), property_ID, GeoUID)) %>% 
  group_by(GeoUID) %>% 
  summarize(FREH = sum(FREH_3))

GH_WD <- 
  GH %>% 
  filter(date %in% as.Date(c("2019-12-31", "2020-12-31"))) %>% 
  mutate(geometry = st_centroid(geometry)) %>% 
  st_join(WD) %>% 
  st_drop_geometry() %>% 
  group_by(ward, date) %>% 
  summarize(GH = sum(housing_units, na.rm = TRUE)) %>% 
  as_tibble() %>% 
  mutate(date = if_else(date >= start_2020, 2020, 2019))

GH_DA <- 
  GH %>% 
  filter(date == "2020-12-31") %>% 
  mutate(geometry = st_centroid(geometry)) %>% 
  st_join(DA) %>% 
  st_drop_geometry() %>% 
  group_by(GeoUID) %>% 
  summarize(GH = sum(housing_units, na.rm = TRUE)) %>% 
  as_tibble()

FREH_total <- 
  daily %>% 
  filter(date >= "2016-01-01") %>% 
  group_by(date) %>% 
  summarize(across(c(FREH, FREH_3), sum)) %>%
  filter(substr(date, 9, 10) == "01")

GH_total <-
  GH %>%
  st_drop_geometry() %>%
  filter(status != "B") %>% 
  group_by(date) %>%
  summarize(GH = sum(housing_units)) %>%
  mutate(GH = slide_dbl(GH, mean, .before = 29))

housing_loss <-
  FREH_total %>%
  select(date, FREH_3) %>% 
  left_join(GH_total, by = "date") %>%
  rename(`Entire home/apt` = FREH_3, `Private room` = GH) %>%
  pivot_longer(c(`Entire home/apt`, `Private room`), 
               names_to = "Listing type",
               values_to = "Housing units") %>% 
  mutate(`Listing type` = factor(`Listing type`, 
                                 levels = c("Private room", "Entire home/apt")))  

housing_loss_WD <- 
  WD %>% 
  left_join(FREH_WD) %>% 
  left_join(GH_WD) %>% 
  mutate(GH = if_else(is.na(GH), 0L, GH),
         housing_loss_pct = (FREH + GH) / dwellings)

housing_loss_DA <- 
  DA %>% 
  left_join(FREH_DA) %>% 
  left_join(GH_DA) %>% 
  mutate(GH = if_else(is.na(GH), 0L, GH),
         housing_loss_pct = (FREH + GH) / dwellings)

WD_housing_table <- 
  WD %>% 
  st_drop_geometry() %>% 
  left_join(FREH_WD) %>% 
  left_join(GH_WD) %>%
  mutate(GH = if_else(is.na(GH), 0L, GH)) %>% 
  group_by(ward) %>% 
  summarize(
    dwellings = mean(dwellings),
    housing_loss_2019 = sum(FREH[date == 2019]) + sum(GH[date == 2019]),
    housing_loss_2020 = sum(FREH[date == 2020]) + sum(GH[date == 2020]),
    yoy_change = (housing_loss_2020 - housing_loss_2019) / housing_loss_2019,
    housing_loss_pct_2020 = housing_loss_2020 / dwellings)
  
rm(property, daily, GH, CMA, DA, city)

```

```{r exec_summ, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

qload(here("output", "str_processed.qsm"), nthreads = availableCores())

# Total housing loss for 2020
housing_loss_2020 <- 
  {{FREH %>% 
      filter(date == "2020-12-01") %>% 
      pull(FREH_3)} +
      {GH %>% 
          filter(date == "2020-12-01") %>% 
          pull(housing_units) %>% sum()}} %>% 
  round(digit = -1) %>% 
  prettyNum(",")

# Total housing loss for 2019
housing_loss_2019 <- 
  {{FREH %>% 
      filter(date == "2019-12-01") %>% 
      pull(FREH_3)} +
      {GH %>% 
          filter(date == end_2019) %>% 
          pull(housing_units) %>% sum()}} %>% 
  round(digit = -1) %>% 
  prettyNum(",")

# Absolute numbers of change in dedicated STRs from 2019 to 2020
housing_loss_change_absolute <- 
  {{{FREH %>% filter(date == as.Date("2019-12-01")) %>% pull(FREH_3)} +
      {GH %>% filter(date == as.Date("2020-12-01") - years(1)) %>% 
          pull(housing_units) %>% sum()}} -
  {{FREH %>% filter(date == as.Date("2020-12-01")) %>% pull(FREH_3)} +
      {GH %>% filter(date == as.Date("2020-12-01")) %>% pull(housing_units) %>% sum()}}} %>% 
  round(digit = -1) %>% 
  prettyNum(",")

# YOY increase in housing loss
housing_loss_change <- 
  {{{FREH %>% filter(date == as.Date("2020-12-01")) %>% pull(FREH_3)} +
      {GH %>% filter(date == as.Date("2020-12-01")) %>% pull(housing_units) %>% sum()}} /
      {{FREH %>% filter(date == as.Date("2019-12-01")) %>% pull(FREH_3)} +
          {GH %>% filter(date == as.Date("2020-12-01") - years(1)) %>% 
              pull(housing_units) %>% sum()}}} %>% 
  {. - 1} %>% 
  round(3) %>% 
  scales::percent(0.1)

housing_loss_pct_dwelling <- 
  {{{FREH %>% filter(date == "2020-12-01") %>% pull(FREH_3)} +
      {GH %>% filter(date == "2020-12-01") %>% pull(housing_units) %>% sum()}} /
      sum(WD$dwellings)} %>% 
  round(3) %>% 
  scales::percent(0.1)

# Housing loss in Spadina-Fort York, University-Rosedale, and Toronto Centre
housing_loss_pct_WD <- 
  WD %>% 
  left_join(FREH_WD) %>% 
  left_join(GH_WD) %>% 
  filter(date == "2020") %>% 
  mutate(GH = if_else(is.na(GH), 0L, GH),
         housing_loss_per_dwelling = (FREH + GH) / dwellings,
         housing_loss_per_dwelling = round(housing_loss_per_dwelling, 3)) %>% 
  st_drop_geometry() %>% 
  filter(ward %in% c("Spadina-Fort York", "University-Rosedale", "Toronto Centre"))

housing_loss_pct_spadina_2019 <- 
  WD %>% 
  left_join(FREH_WD) %>% 
  left_join(GH_WD) %>% 
  filter(date == "2019") %>% 
  mutate(GH = if_else(is.na(GH), 0L, GH),
         housing_loss_per_dwelling = (FREH + GH) / dwellings,
         housing_loss_per_dwelling = round(housing_loss_per_dwelling, 3)) %>% 
  st_drop_geometry() %>% 
  filter(ward %in% c("Spadina-Fort York")) %>% 
    pull(housing_loss_per_dwelling) %>% 
  scales::percent(0.1)

housing_loss_pct_spadina <- 
  housing_loss_pct_WD %>% 
  filter(ward == "Spadina-Fort York") %>% 
  pull(housing_loss_per_dwelling) %>% 
  scales::percent(0.1)

housing_loss_pct_university <- 
  housing_loss_pct_WD %>% 
  filter(ward == "University-Rosedale") %>% 
  pull(housing_loss_per_dwelling) %>% 
  scales::percent(0.1)

housing_loss_pct_centre <- 
  housing_loss_pct_WD %>% 
  filter(ward == "Toronto Centre") %>% 
  pull(housing_loss_per_dwelling) %>% 
  scales::percent(0.1)


rm(property, daily, GH)

```

**Short-term rentals removed `r housing_loss_2020` housing units from Toronto’s long-term market in 2020 — a decrease of `r housing_loss_change` from 2019. The Covid-19 pandemic has reduced the number of dedicated STRs in Toronto by `r housing_loss_change_absolute` units compared to the year before. Almost two thirds of all entire-home listings and four in ten private-room listings are dedicated STRs. In the Spadina-Fort York area, `r housing_loss_pct_spadina` of all housing has been converted to dedicated STRs, while the figure is `r housing_loss_pct_university` in University-Rosedale and `r housing_loss_pct_centre` in Toronto Centre. Between 2015 and 2019, STRs have been responsible for a 2.1% increase in average monthly rent, and the average renter household in Toronto has paid an additional $1,390 in rent because of the impact of STRs.**

<br>

## STR-induced housing loss

Toronto’s housing market has been under considerable stress in the past years, with housing prices and rents rising, and rental vacancy rates falling. These are symptoms of a market where the supply of housing is insufficient to meet demand. One possible explanation for both the insufficient supply and elevated demand for housing in Toronto is the growth in short-term rentals. Tourists are now able to compete with residents for housing—adding demand to the local housing market—while landlords are now able to shift their properties out of the conventional housing market to become dedicated STRs—reducing the supply of conventional housing. Research has found that renting a housing unit on the STR market frequently offers landlords greater potential revenue than conventional leases (Wachsmuth & Weisler 2018), especially in transit-accessible neighborhoods (Deboosere et al. 2019). Multiple studies have also found that Airbnb and other STR platforms increase housing costs (Barron, Kung, & Proserpio 2017; Horn & Merante 2017; Garcia-Lopez et al. 2019).

One of the major considerations when gauging the impacts of short-term rentals (STRs) on a city, therefore, is the extent to which STRs are removing long-term housing from the market. This process can occur either directly, where tenants of a unit are evicted or not replaced at the end of a lease and the unit is converted to a STR, or indirectly by absorbing new construction or investment properties which otherwise would have gone onto the long-term market. To obtain the exact number of units that have been occupied as STRs, landlords or units would need to be individually surveyed, which is infeasible because STR hosts are mostly anonymous on major STR platforms such as Airbnb and VRBO. Instead, we use the daily activity of listings, alongside structural characteristics such as listing type and location, to estimate which listings are operating as dedicated STRs and are therefore not available as conventional long-term housing.

_Frequently Rented Entire-Home (FREH) listings:_ The number of frequently-rented units is one way to estimate STR-induced housing loss. If a STR is available for reservations the majority of the year and receives many bookings, it is reasonable to assume that it is not serving as an individual’s principal residence at the same time. Along these lines, we define frequently rented entire-home (FREH) listings as entire-home listings which were available on Airbnb or VRBO the majority of the year (at least 183 nights) and were booked a minimum of 90 nights. We then apply a statistical model (described in the appendix) to the FREH data in order to generate an estimate of FREH activity based on three months of listing activity. This allows us to detect listings which are operating in a full-time manner but have not yet been listed for an entire year, and allows us to account for relatively short-term changes in market conditions.

_Ghost hostels:_ In addition to FREH listings, it is possible that entire housing units have been subdivided into multiple private-room listings, each of which appearing to be a spare bedroom or the like, while actually collectively representing an apartment removed from the long-term housing market. We call these clusters of private-room listings “ghost hostels”, building on the advocacy group Fairbnb.ca’s term “ghost hotels”—multiple FREH listings located in a single building, collectively serving as de facto hotels instead of long-term housing (Wieditz 2017). We detect ghost hostels by finding clusters of three or more private-room listings operated by a single host, whose reported locations are close enough to each other that they are likely to have originated in the same actual housing unit. (Airbnb and VRBO obfuscate listing locations by shifting them randomly up to 200 m.)

```{r para_1, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

qload(here("output", "str_processed.qsm"), nthreads = availableCores())

# FREH in 2020-10
FREH_2020 <- 
  FREH %>% 
  filter(date == "2020-12-01") %>% 
  pull(FREH_3) %>% 
  round(digit = -1) %>% 
  prettyNum(",")

# GH in 2020-10
GH_2020 <- 
  GH %>% 
  filter(date == "2020-12-01") %>% 
  pull(housing_units) %>% 
  sum() %>% 
  round(digit = -1) %>% 
  prettyNum(",")

# Change in listings
active_yoy_change <- 
  daily %>% 
  filter(housing, status != "B", date >= start_2020 - years(1), 
         date <= end_2020) %>% 
  group_by(year_2020 = date >= start_2020) %>% 
  summarize(active_listings = n() / 365,
            revenue = sum(price[status == "R"]), .groups = "drop") %>% 
  summarize(across(c(active_listings, revenue), ~{(.x[2] - .x[1]) / .x[1]}),
            .groups = "drop")

# YOY listing change
active_yoy_listings <- 
  active_yoy_change$active_listings %>% 
  {. * -1} %>% 
  scales::percent(accuracy = 0.1)

# Total housing loss for 2020
housing_loss_2020 <- 
  {{FREH %>% 
      filter(date == "2020-12-01") %>% 
      pull(FREH_3)} +
    {GH %>% 
        filter(date == "2020-12-01") %>% 
        pull(housing_units) %>% 
        sum()}} %>% 
  round(digit = -1) %>% 
  prettyNum(",")

rm(property, daily, GH)

```


``` {r make_fig_2_1}

figure_2_1_fun <- function(regular = "", condensed = "") {
  
housing_loss %>% 
  ggplot(aes(date, `Housing units`, fill = `Listing type`)) +
  geom_col(lwd = 0) +
  annotate("segment", x = key_date_covid, xend = key_date_covid,
           y = 0, yend = Inf, alpha = 0.3) +
  annotate("curve", x = as.Date("2019-08-01"), 
           xend = key_date_covid - days(10),
           y = 7000, yend = 6700, curvature = -.2, lwd = 0.25,
           arrow = arrow(length = unit(0.05, "inches"))) +
  annotate("text", x = as.Date("2019-05-01"), y = 7000,
           label = "COVID-19 \nAirbnb's response", family = condensed) +
  annotate("segment", x = key_date_regulations, xend = key_date_regulations,
           y = 0, yend = Inf, alpha = 0.3) +
  annotate("curve", x = as.Date("2020-06-01"), 
           xend = key_date_regulations - days(10),
           y = 7200, yend = 7100, curvature = -.2, lwd = 0.25,
           arrow = arrow(length = unit(0.05, "inches"))) +
  annotate("text", x = as.Date("2020-06-01"), y = 7100,
           label = "Regulations", family = condensed) +
  scale_fill_manual(values = col_palette[c(2, 4)]) +
  scale_x_date(name = NULL, limits = c(as.Date("2016-10-01"), NA)) +
  scale_y_continuous(name = NULL, label = scales::comma) +
  theme_minimal() +
  theme(legend.position = "bottom", 
        panel.grid.minor.x = element_blank(),
        text = element_text(face = "plain", family = regular),
        legend.title = element_text(face = "bold", family = regular, 
                                    size = 10),
        legend.text = element_text( size = 10, family = regular))
  
}

if (build_figures) {
  ggsave(here("output", "figures", "figure_2_1.pdf"), 
         plot = figure_2_1_fun("Futura", "Futura Condensed"), 
         width = 8, height = 5, units = "in", useDingbats = FALSE)
  
  extrafont::embed_fonts(here("output", "figures", "figure_2_1.pdf"))
}

```

At the end of 2020, there were `r FREH_2020` FREH listings in the City of Toronto, and `r GH_2020` more housing units which were operating as ghost hostels. In total, therefore, short-term rentals removed `r housing_loss_2020` housing units from Toronto’s long-term market last year (Figure \@ref(fig:fig-2-1)). Notably, while the number of active daily listings declined by `r active_yoy_listings` over 2020, the number of housing units which STRs took off of Toronto’s housing market decreased at a much sharper rate. There were `r housing_loss_change` fewer dedicated STRs in 2020 compared to 2019, from `r housing_loss_2019` to `r housing_loss_2020`. This high decline rate in commercially-operated STRs reflects that the pandemic has disproportionately affected commercial operators in the City.


``` {r fig-2-1, include = TRUE, fig.cap = '(ref:fig-2-1)', fig.align = "center"}

figure_2_1_fun()

```

(ref:fig-2-1) _Housing units converted to dedicated STRs in the City of Toronto (monthly average)_

```{r para_2, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

qload(here("output", "str_processed.qsm"), nthreads = availableCores())

# End-of-year summaries
housing_loss_summaries <- 
  daily %>% 
  filter(housing, status != "B", listing_type %in% c("Entire home/apt",
                                                     "Private room")) %>% 
  group_by(date) %>% 
  summarize(eh = mean(FREH_3[listing_type == "Entire home/apt"] > 0.5),
            pr = mean(GH[listing_type == "Private room"])) %>% 
  mutate(across(c(eh, pr), slide_dbl, mean, .before = 30)) %>% 
  filter(date %in% as.Date(c("2016-12-31", "2019-12-31", "2020-12-31")))

eh_housing_loss <- 
  housing_loss_summaries %>% 
  filter(date == "2020-12-31") %>% 
  pull(eh) %>% 
  scales::percent(0.1)

pr_housing_loss <- 
  housing_loss_summaries %>% 
  filter(date == "2020-12-31") %>% 
  pull(pr) %>% 
  scales::percent(0.1)

eh_housing_loss_2019 <- 
  housing_loss_summaries %>% 
  filter(date == "2020-12-31") %>% 
  pull(eh) %>% 
  scales::percent(0.1)

pr_housing_loss_2019 <- 
  housing_loss_summaries %>% 
  filter(date == "2020-12-31") %>% 
  pull(pr) %>% 
  scales::percent(0.1)

eh_housing_loss_2016 <- 
  housing_loss_summaries %>% 
  filter(date == "2016-12-31") %>% 
  pull(eh) %>%
  scales::percent(0.1)

pr_housing_loss_2016 <- 
  housing_loss_summaries %>% 
  filter(date == "2016-12-31") %>% 
  pull(pr) %>% 
  scales::percent(0.1)

# Table for figure
housing_loss_share <- 
  daily %>% 
  filter(housing, status != "B", listing_type %in% c("Entire home/apt",
                                                     "Private room")) %>% 
  group_by(date) %>% 
  summarize(
    `Entire home/apt` = mean(FREH_3[listing_type == "Entire home/apt"] > 0.5),
    `Private room` = mean(GH[listing_type == "Private room"])) %>% 
  filter(date >= "2016-10-01") %>% 
  pivot_longer(-date, names_to = "Listing type", 
               values_to = "housing_loss_pct") %>% 
  group_by(`Listing type`) %>% 
  mutate(housing_loss_pct = slide_dbl(housing_loss_pct, mean, .before = 13))

rm(property, daily, GH)

```

``` {r make_fig_2_2}

figure_2_2_fun <- function(regular = "", condensed = "") {
  
  housing_loss_share %>% 
    ggplot(aes(date, housing_loss_pct, colour = `Listing type`)) +
    geom_line(lwd = 1) +
    annotate("segment", x = key_date_covid, xend = key_date_covid,
             y = 0, yend = Inf, alpha = 0.3) +
    annotate("segment", x = key_date_regulations, xend = key_date_regulations,
             y = 0, yend = Inf, alpha = 0.3) +
    scale_x_date(name=NULL) +
    scale_y_continuous(name = NULL, labels = scales::percent) +
    scale_colour_manual(values = col_palette[c(2, 4)]) +
    theme_minimal() +
    theme(legend.position = "bottom",
          panel.grid.minor.x = element_blank(),
          text = element_text(family = regular, face = "plain"),
          legend.title = element_text(family = regular, face = "bold"),
          legend.text = element_text(family = regular))
  
}

if (build_figures) {
  ggsave(here("output", "figures", "figure_2_2.pdf"), 
         plot = figure_2_2_fun("Futura", "Futura Condensed"), 
         width = 8, height = 5, units = "in", useDingbats = FALSE)
  
  extrafont::embed_fonts(here("output", "figures", "figure_2_2.pdf"))
}

```

Combined with the falling numbers of listings contributing to housing loss is the decreasing proportion of STRs out of all active listings which are either FREH or ghosts hostels in 2020. At the end of 2020 only a quarter (`r eh_housing_loss`) of entire-home listings and less than half of (`r pr_housing_loss`) private-room listings were taking housing off the market in Toronto (Figure \@ref(fig:fig-2-2)). In 2019, two thirds of entire-home listings (`r eh_housing_loss_2019`) and half (`r pr_housing_loss_2019`) of the private-room listings contributed to housing loss. It is interesting to note that the proportion of entire-home listings contributing to housing loss has decreased to even lower levels than that of 2016 (`r eh_housing_loss_2019`) thanks to the pandemic.

``` {r fig-2-2, include = TRUE, fig.cap = '(ref:fig-2-2)', fig.align = "center"}

figure_2_2_fun()

```

(ref:fig-2-2) _The percentage of active STR listings contributing to housing loss each day in Toronto (14-day average)_

Table \@ref(tab:tab-2-1) summarizes STR-induced housing loss patterns by neighbourhood. It demonstrates that the City-wide trend of shrinking housing loss due to the pandemic holds in all neighbourhoods. Also, year-over-year decline of housing loss has been evenly distributed between virtually all wards, with an average of -59.3%. The following section takes a closer look at the impact of Covid-19 on frequently rented entire-home listings.

``` {r tab-2-1, include = TRUE}

library(kableExtra)

WD_housing_table %>% 
  summarize(
    dwellings = sum(dwellings),
    housing_loss_2019 = sum(housing_loss_2019),
    housing_loss_2020 = sum(housing_loss_2020),
    yoy_change = (housing_loss_2020 - housing_loss_2019) / housing_loss_2019,
    housing_loss_pct_2019 = housing_loss_2020 / dwellings) %>% 
  mutate(ward = "City of Toronto") %>% 
  bind_rows(WD_housing_table) %>% 
  relocate(ward) %>% 
  select(-dwellings) %>% 
  filter(housing_loss_2020 > 50) %>% 
  mutate(across(c(housing_loss_2019, housing_loss_2020), round, -1)) %>% 
  mutate(across(c(yoy_change, housing_loss_pct_2019), round, 4)) %>% 
  arrange(desc(housing_loss_2020)) %>%
  rename(Ward = ward,
         `Housing loss (2019)` = housing_loss_2020,
         `Housing loss (2018)` = housing_loss_2019,
         `Annual growth (%)` = yoy_change,
         `% of housing lost (2019)` = housing_loss_pct_2019) %>% 
  kbl(caption = paste0("STR-induced housing loss by ward in the City", 
                       "of Toronto (for boroughs with at least 50 housing ",
                       "units lost in 2019)"),
      align = "lrrrr") %>%
  kable_styling(latex_options = "scale_down") %>%
  row_spec(1, background = paste0(col_palette[2], "50"))

```

```{r para_3, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

qload(here("output", "cmhc.qsm"), nthreads = availableCores())

# Vacancy rate downtown (Toronto (Central))
downtown_spadina <- 
  annual_vacancy %>% 
  filter(dwelling_type == "Total",
         bedroom == "Total",
         zone %in% c(1),
         !is.na(vacancy)) %>% 
  group_by(zone) %>% 
  filter(date == max(date)) %>% 
  ungroup() %>% 
  pull(vacancy) %>% 
  scales::percent(0.1)

```

``` {r make_fig_2_3}

figure_2_3_fun <- function(regular = "", condensed = "") {
  
  left_map <- 
    housing_loss_WD %>% 
    filter(date=="2020") %>% 
    ggplot() +
    geom_sf(data = province, colour = "transparent", fill = "grey93") +
    geom_sf(aes(fill = housing_loss_pct), colour = "white") +
    scale_fill_gradientn(colors = col_palette[c(3, 4, 5)], 
                         na.value = "grey80",
                         limits = c(0, 0.03), oob = scales::squish, 
                         labels = scales::percent)  +
    guides(fill = guide_colourbar(title = "% housing\nlost to STR",
                                  title.vjust = 1)) + 
    gg_bbox(housing_loss_DA) +
    theme_void() +
    theme(text = element_text(family = regular, face = "plain"),
          legend.title = element_text(family = regular, face = "bold", 
                                      size = 7),
          legend.title.align = 0.9,
          legend.text = element_text(family = regular, size = 5))
  
  right_map <- 
    housing_loss_DA %>% 
    ggplot() +
    geom_sf(data = province, colour = "transparent", fill = "grey93") +
    geom_sf(aes(fill = housing_loss_pct), colour = "transparent") +
    scale_fill_gradientn(colors = col_palette[c(3, 4, 5)], 
                         na.value = "grey80",
                         limits = c(0, 0.05), oob = scales::squish, 
                         labels = scales::percent)  +
    guides(fill = guide_colourbar(title = "% housing\nlost to STR",
                                  title.vjust = 1)) + 
    gg_bbox(housing_loss_DA) +
    theme_void() +
    theme(text = element_text(family = regular, face = "plain"),
          legend.title = element_text(family = regular, face = "bold", 
                                      size = 7),
          legend.title.align = 0.9,
          legend.text = element_text(family = regular, size = 5))
  
  left_map + right_map + plot_layout(guides = 'collect') & 
    theme(legend.position = "bottom")

}

if (build_figures) {
  ggsave(here("output", "figures", "figure_2_3.pdf"), 
         plot = figure_2_3_fun("Futura", "Futura Condensed"), 
         width = 8, height = 4.2, units = "in", useDingbats = FALSE)
  
  extrafont::embed_fonts(here("output", "figures", "figure_2_3.pdf"))
}

```

The `r housing_loss_2020` housing units taken off of Toronto’s housing market in 2020 are `r housing_loss_pct_dwelling` of the total amount of housing in the city, and this housing loss has been concentrated in small parts of the city. This is more or less the same in relative to the total housing stock: Figure \@ref(fig:fig-2-3) shows the proportion of each ward or dissemination area’s housing stock which was operated as a dedicated short-term rental as of the end of 2020. The maps show a tale of two cities: in most of Toronto, there are relatively few dedicated STRs, while in the center of the city as well as a few transit corridors, they are ubiquitous. In the Spadina-Fort York ward, `r housing_loss_pct_spadina` of all housing units have been converted to dedicated STRs. The figure is `r housing_loss_pct_university` for University-Rosedale and `r housing_loss_pct_centre` for Toronto Centre. In the Downtown CMHC zone, the rental vacancy rate was `r downtown_spadina` in 2019, while the number of STRs contributing to housing loss in the Spadina-Fort York in 2019 was `r housing_loss_pct_spadina_2019`. This means that there were slightly more dedicated STRs in this neighbourhood than there were vacant apartments for rent. (2020 rental vacancy rates numbers were not yet available at the time of analysis.)

``` {r fig-2-3, include = TRUE, fig.cap = '(ref:fig-2-3)', fig.align = "center"}

figure_2_3_fun()

```

(ref:fig-2-3) _The percentage of housing units converted to dedicated STRs in the City of Toronto, by ward (L) and dissemination area (R)_

## Covid’s impact on frequently rented entire-home listings

```{r FREH_outcomes, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

qload(here("output", "str_processed.qsm"), nthreads = availableCores())

FREH_2020 <- 
  daily %>% 
  filter(housing, date >= start_2020, date <= "2020-02-29", FREH_3 > 0.5) %>% 
  pull(property_ID) %>% 
  unique() %>% 
  {filter(property, property_ID %in% .)} %>% 
  st_drop_geometry()

non_FREH_2020 <- 
  property %>% 
  st_drop_geometry() %>% 
  filter(property_ID %in% {daily %>% 
      filter(housing, date >= start_2020, date <= "2020-02-29", 
             status != "B") %>% 
      pull(property_ID) %>% 
      unique()}, !property_ID %in% FREH_2020$property_ID)

FREH_2019 <- 
  daily %>% 
  filter(housing, date >= "2019-01-01", date <= "2019-02-28", FREH_3 > 0.5) %>% 
  pull(property_ID) %>% 
  unique() %>% 
  {filter(property, property_ID %in% .)} %>% 
  st_drop_geometry()

non_FREH_2019 <- 
  property %>% 
  st_drop_geometry() %>% 
  filter(property_ID %in% {daily %>% 
      filter(housing, date >= "2019-01-01", date <= "2019-02-28", 
             status != "B") %>% 
      pull(property_ID) %>% 
      unique()}, !property_ID %in% FREH_2019$property_ID)

total_2020 <- 
  tibble(
    group = c("FREH", "non-FREH"),
    year = 2020,
    variable = "total listings",
    value = c(nrow(FREH_2020), nrow(non_FREH_2020))
  )

total_2019 <- 
  tibble(
    group = c("FREH", "non-FREH"),
    year = 2019,
    variable = "total listings",
    value = c(nrow(FREH_2019), nrow(non_FREH_2019))
  )

deactivated_2020 <- 
  tibble(
    group = c("FREH", "non-FREH"),
    year = 2020,
    variable = "deactivated",
    value = c({
      FREH_2020 %>% 
        summarize(total = sum(scraped <= "2020-12-31")) %>% 
        pull(total)}, {
          non_FREH_2020 %>% 
            summarize(total = sum(scraped <= "2020-12-31")) %>% 
            pull(total)})
  )

deactivated_2019 <- 
  tibble(
    group = c("FREH", "non-FREH"),
    year = 2019,
    variable = "deactivated",
    value = c({
      FREH_2019 %>% 
        summarize(total = sum(scraped <= "2019-12-31")) %>% 
        pull(total)}, {
          non_FREH_2019 %>% 
            summarize(total = sum(scraped <= "2019-12-31")) %>% 
            pull(total)})
  )

blocked_PIDs_2020 <- 
  daily %>% 
  filter(housing, date >= "2020-12-01", date <= "2020-12-31") %>% 
  group_by(property_ID) %>% 
  filter(mean(status == "B") == 1) %>% 
  pull(property_ID) %>% 
  unique()

blocked_PIDs_2019 <- 
  daily %>% 
  filter(housing, date >= "2019-12-01", date <= "2019-12-31") %>% 
  group_by(property_ID) %>% 
  filter(mean(status == "B") == 1) %>% 
  pull(property_ID) %>% 
  unique()

blocked_2020 <-
  tibble(
    group = c("FREH", "non-FREH"),
    year = 2020,
    variable = "blocked",
    value = c(nrow(filter(FREH_2020, property_ID %in% blocked_PIDs_2020)),
              nrow(filter(non_FREH_2020, property_ID %in% blocked_PIDs_2020)))
  )

blocked_2019 <-
  tibble(
    group = c("FREH", "non-FREH"),
    year = 2019,
    variable = "blocked",
    value = c(nrow(filter(FREH_2019, property_ID %in% blocked_PIDs_2019)),
              nrow(filter(non_FREH_2019, property_ID %in% blocked_PIDs_2019)))
  )

active_2020 <- 
  tibble(
    group = c("FREH", "non-FREH"),
    year = 2020,
    variable = "active",
    value = c(
      nrow(FREH_2020) - deactivated_2020[1,]$value - blocked_2020[1,]$value,
      nrow(non_FREH_2020) - deactivated_2020[2,]$value - blocked_2020[2,]$value)
  )

active_2019 <- 
  tibble(
    group = c("FREH", "non-FREH"),
    year = 2019,
    variable = "active",
    value = c(
      nrow(FREH_2019) - deactivated_2019[1,]$value - blocked_2019[1,]$value,
      nrow(non_FREH_2019) - deactivated_2019[2,]$value - blocked_2019[2,]$value)
  )

comparison_df <- 
  bind_rows(total_2020, total_2019, deactivated_2020, deactivated_2019,
            blocked_2020, blocked_2019, active_2020, active_2019)

# Figure parameters
n_segs <- 1000
offset <- 1000
max_y <- comparison_df %>% filter(value == max(value)) %>% 
  mutate(value = value + 2 * offset) %>% pull(value)

fig_polys <- 
  comparison_df %>% 
  group_by(group, year, variable) %>%
  summarize(
    x = c(3, 5) - (variable == "total listings") * 3,
    ymax = rep(value, 2)) %>% 
  ungroup() %>% 
  arrange(desc(variable)) %>% 
  group_by(group, year) %>% 
  mutate(
    ymin = case_when(
      variable == "total listings" ~ offset,
      variable == "active"         ~ 0,
      variable == "blocked"        ~ ymax[variable == "active"][1] + offset,
      variable == "deactivated"    ~ ymax[variable == "blocked"][1] + 
        ymax[variable == "active"][1] + offset * 2
    ),
    ymax = ymax + ymin, .before = ymax) %>% 
  mutate(adjustment = max_y / 2 - mean(c(ymax[variable == "total listings"], 
                                         ymin[variable == "total listings"])),
         ymin = ymin + adjustment,
         ymax = ymax + adjustment) %>% 
  ungroup()

fig_segments <-
  fig_polys %>% 
  filter(variable != "total listings") %>% 
  mutate(orientation = case_when(
    variable == "deactivated" ~ -1,
    variable == "blocked"     ~ 0,
    variable == "active"      ~ 1)) %>% 
  group_by(group, year, variable) %>% 
  mutate(across(c(ymin, ymax), ~{.x + c(offset, 0) * orientation})) %>% 
  ungroup() %>% 
  select(-orientation) %>% 
  mutate(x = (x + 1) / 2) %>% 
  group_by(group, year) %>% 
  mutate(ramp = 1:6) %>% 
  group_by(group, year, variable) %>% 
  mutate(ymin_slope = diff(ymin) / diff(x),
         ymin_intercept = ymin[1] - ymin_slope * x[1],
         ymax_slope = diff(ymax) / diff(x),
         ymax_intercept = ymax[1] - ymax_slope * x[1]) %>% 
  summarize(
    x = seq(min(x), max(x), length.out = n_segs),
    xend = x,
    y = x * ymin_slope[1] + ymin_intercept[1],
    yend = xend * ymax_slope[1] + ymax_intercept[1],
    ramp = x - min(x) + min(ramp)
  ) %>% 
  ungroup()

fig_labels <-
  fig_polys %>% 
  group_by(group, year, variable) %>% 
  summarize(x = mean(x),
            y = mean(c(ymin, ymax)),
            label = mean(ymax) - mean(ymin)) %>% 
  ungroup() %>% 
  group_by(group, year) %>% 
  mutate(label = case_when(
    variable == "total listings" ~ paste0(label, " ", variable, "\nin Jan/Feb"),
    variable == "active" ~ paste0(label, " (", 
                                  round(label/sum(label) * 200, 1), "%) ", 
                                  variable, "\nin December"),
    TRUE ~ paste0(label, " (", round(label/sum(label) * 200, 1), "%) ",
                  variable))) %>% 
  ungroup()

# Facts about housing loss
GH_total <- 
  GH %>% 
  st_drop_geometry() %>% 
  filter(status != "B") %>% 
  group_by(date) %>% 
  summarize(GH_units = sum(housing_units))

housing_loss_summary <- 
  daily %>% 
  group_by(date) %>% 
  summarize(FREH = sum(FREH_3)) %>% 
  left_join(GH_total) %>% 
  mutate(housing_loss = FREH + GH_units) %>% 
  filter(date >= key_date_regulations) %>% 
  filter(housing_loss == max(housing_loss, na.rm = TRUE)) %>% 
  mutate(across(-date, round, -1))

peak_housing_loss <- 
  prettyNum(housing_loss_summary$housing_loss, ",")

peak_FREH <- 
  prettyNum(housing_loss_summary$FREH, ",")

peak_GH <-
  prettyNum(housing_loss_summary$GH_units, ",")

oct_2020_housing_loss <- 
  daily %>% 
  group_by(date) %>% 
  summarize(FREH = sum(FREH_3)) %>% 
  left_join(GH_total) %>% 
  mutate(housing_loss = FREH + GH_units) %>% 
  filter(date == "2020-12-01") %>% 
  mutate(across(-date, round, -1)) %>% 
  pull(housing_loss) %>% 
  prettyNum(",")

#' FREH in either Jan or Feb 2020
FREH_in_jan_feb <- 
  daily %>% 
  filter(housing, date >= "2020-01-01", date <= "2020-02-29", FREH_3 > 0.5) %>% 
  pull(property_ID) %>% 
  unique()

jan_feb_FREH_total <- 
  length(FREH_in_jan_feb) %>% 
  round(-1) %>% 
  prettyNum(",")

jan_feb_FREH_deactivated <- 
  property %>% 
  st_drop_geometry() %>% 
  filter(property_ID %in% FREH_in_jan_feb) %>% 
  summarize(total = round(sum(scraped <= "2020-12-31"), -1),
            pct = round(mean(scraped <= "2020-12-31"), 3)) %>% 
  pull(total) %>% 
  prettyNum(",")

jan_feb_FREH_deactivated_pct <- 
  property %>% 
  st_drop_geometry() %>% 
  filter(property_ID %in% FREH_in_jan_feb) %>% 
  summarize(total = round(sum(scraped <= "2020-12-31"), -1),
            pct = round(mean(scraped <= "2020-12-31"), 3)) %>% 
  pull(pct) %>% 
  scales::percent(0.1)

FREH_in_jan_feb_2019 <- 
  daily %>% 
  filter(housing, date >= "2019-01-01", date <= "2019-02-28", FREH_3 > 0.5) %>% 
  pull(property_ID) %>% 
  unique()

jan_feb_FREH_deactivated_pct_2019 <- 
  property %>% 
  st_drop_geometry() %>% 
  filter(property_ID %in% FREH_in_jan_feb_2019) %>% 
  summarize(pct = mean(scraped <= "2019-12-31")) %>% 
  pull(pct) %>% 
  scales::percent(0.1)

jan_feb_non_FREH_deactivated_pct <- 
  property %>% 
  st_drop_geometry() %>% 
  filter(property_ID %in% {daily %>% 
      filter(housing, date >= "2020-01-01", date <= "2020-02-29", 
             status != "B") %>% 
      pull(property_ID) %>% 
      unique()}, !property_ID %in% FREH_in_jan_feb) %>% 
  summarize(pct = mean(scraped <= "2020-12-31")) %>% 
  pull(pct) %>% 
  scales::percent(0.1)

jan_feb_non_FREH_deactivated_pct_2019 <- 
  property %>% 
  st_drop_geometry() %>% 
  filter(property_ID %in% {daily %>% 
      filter(housing, date >= "2019-01-01", date <= "2019-02-28", 
             status != "B") %>% 
      pull(property_ID) %>% 
      unique()}, !property_ID %in% FREH_in_jan_feb_2019) %>% 
  summarize(pct = mean(scraped <= "2019-12-31")) %>% 
  pull(pct) %>% 
  scales::percent(0.1)

# For final figure
monthly_reservation_trajectories <- 
  daily %>% 
  filter(housing, status == "R", date >= "2019-01-01", 
         listing_type == "Entire home/apt") %>% 
  mutate(FREH_feb = if_else(property_ID %in% FREH_in_jan_feb, TRUE, FALSE)) %>% 
  tsibble::as_tsibble(index = date, key = property_ID) %>% 
  tsibble::group_by_key() %>% 
  tsibble::index_by(yearmon = tsibble::yearmonth(date)) %>% 
  summarize(n = sum(status == "R"),
            FREH_feb = as.logical(prod(FREH_feb))) %>% 
  group_by(FREH_feb) %>% 
  tsibble::index_by(yearmon) %>% 
  summarize(n = mean(n)) %>% 
  arrange(yearmon, FREH_feb)

daily_reservation_trajectories <- 
  daily %>% 
  filter(housing, status == "R", date >= "2019-01-01", 
         listing_type == "Entire home/apt") %>% 
  mutate(FREH_feb = if_else(property_ID %in% FREH_in_jan_feb, TRUE, FALSE)) %>% 
  count(date, FREH_feb)

jan_feb_FREH_active <- 
  property %>% 
  filter(property_ID %in% FREH_in_jan_feb, scraped > "2020-12-31") %>% 
  nrow() %>% 
  round(-1) %>% 
  prettyNum(",")

jan_feb_FREH_blocked_all <- 
  daily %>% 
  filter(housing, date >= "2020-12-01", date <= "2020-12-31") %>% 
  group_by(property_ID) %>% 
  filter(mean(status == "B") == 1) %>% 
  pull(property_ID) %>% 
  unique() %>% 
  {filter(property, property_ID %in% ., property_ID %in% FREH_in_jan_feb)} %>% 
  nrow() %>% 
  round(-1) %>% 
  prettyNum(",")

jan_feb_FREH_blocked_all_pct <- 
  {daily %>% 
  filter(housing, date >= "2020-12-01", date <= "2020-12-31") %>% 
  group_by(property_ID) %>% 
  filter(mean(status == "B") == 1) %>% 
  pull(property_ID) %>% 
  unique() %>% 
  {filter(property, property_ID %in% ., property_ID %in% FREH_in_jan_feb)} %>% 
  nrow() %>% 
  `/`(property %>% 
        filter(property_ID %in% FREH_in_jan_feb, scraped > "2020-12-31") %>% 
        nrow())} %>% 
  scales::percent(0.1)

jan_feb_FREH_blocked_maj <- 
  daily %>% 
  filter(housing, date >= "2020-12-01", date <= "2020-12-31") %>% 
  group_by(property_ID) %>% 
  filter(mean(status == "B") > 0.5) %>% 
  pull(property_ID) %>% 
  unique() %>% 
  {filter(property, property_ID %in% ., property_ID %in% FREH_in_jan_feb)} %>% 
  nrow() %>% 
  round(-1) %>% 
  prettyNum(",")

jan_feb_FREH_blocked_maj_pct <- 
  {daily %>% 
    filter(housing, date >= "2020-12-01", date <= "2020-12-31") %>% 
    group_by(property_ID) %>% 
    filter(mean(status == "B") > 0.5) %>% 
    pull(property_ID) %>% 
    unique() %>% 
    {filter(property, property_ID %in% ., property_ID %in% FREH_in_jan_feb)} %>% 
    nrow() %>% 
    `/`(property %>% 
          filter(property_ID %in% FREH_in_jan_feb, scraped > "2020-12-31") %>% 
          nrow())} %>% 
  scales::percent(0.1)

jan_feb_FREH_blocked_all_pct_2019 <- 
  {daily %>% 
    filter(housing, date >= "2019-12-01", date <= "2019-12-31") %>% 
    group_by(property_ID) %>% 
    filter(mean(status == "B") == 1) %>% 
    pull(property_ID) %>% 
    unique() %>% 
    {filter(property, property_ID %in% ., 
            property_ID %in% FREH_in_jan_feb_2019)} %>% 
    nrow() %>% 
    `/`(property %>% 
          filter(property_ID %in% FREH_in_jan_feb_2019, 
                 scraped > "2019-12-31") %>% 
          nrow())} %>% 
  scales::percent(0.1)

jan_feb_FREH_blocked_maj_pct_2019 <- 
  {daily %>% 
    filter(housing, date >= "2019-12-01", date <= "2019-12-31") %>% 
    group_by(property_ID) %>% 
    filter(mean(status == "B") > 0.5) %>% 
    pull(property_ID) %>% 
    unique() %>% 
    {filter(property, property_ID %in% ., 
            property_ID %in% FREH_in_jan_feb_2019)} %>% 
    nrow() %>% 
    `/`(property %>% 
          filter(property_ID %in% FREH_in_jan_feb_2019, 
                 scraped > "2019-12-31") %>% 
          nrow())} %>% 
  scales::percent(0.1)

feb_2020_FREH_pct <- 
  daily %>% 
  filter(housing, date >= "2020-02-01", date <= "2020-02-29", status == "R") %>% 
  summarize(pct = mean(property_ID %in% FREH_in_jan_feb)) %>% 
  pull(pct) %>% 
  scales::percent(0.1)

rm(property, daily, GH)

```

According to the FREH model, the number of housing units in Toronto lost due to commercial STRs reached its all-time peak (`r peak_housing_loss`) at the beginning of 2020. Most of these (`r peak_FREH`) were FREH listings, with the remainder (`r peak_GH`) being ghost hostels—clusters of private-room listings operated out of a single housing unit. As of December 2020, the number of FREH listings had dropped to its lowest amount since we began tracking it in 2016, with just `r oct_2020_housing_loss` listings displaying availability and reservations consistent with historical patterns of full-time STR activity in Vancouver.

One possibility is that these formerly FREH listings are no longer operating as STRs, either permanently (the listings were deactivated) or temporarily (the listings were blocked from receiving reservations). Under this possibility the listings may have been returned to the longer-term rental market. Another possibility is that the listings have remained open for business as short-term rentals, but the dramatically decreased demand for tourist accommodations means that the listings have not received the level of reservations which our model expects in order to classify the listings as frequently rented.

``` {r make_fig_2_4}

figure_2_4_fun <- function(regular = "", condensed = "") {
  
  fig_polys %>% 
    ggplot() +
    geom_ribbon(aes(x = x, ymin = ymin, ymax = ymax, fill = variable)) +
    geom_segment(aes(x = x, xend = xend, y = y, yend = yend, group = variable,
                     colour = ramp), data = fig_segments, inherit.aes = FALSE) +
    geom_text(aes(x, y, group = variable, label = label), data = fig_labels,
              colour = "white", size = 2, family = regular) + 
    scale_colour_gradientn(colours = col_palette[c(5, 3, 5, 2, 5, 1)]) +
    scale_fill_manual(values = c("total listings" = col_palette[5], 
                                 "active" = col_palette[1],
                                 "blocked" = col_palette[2],
                                 "deactivated" = col_palette[3])) +
    facet_grid(rows = vars(group), cols = vars(year), switch = "y",
               scales = "free_y", space = "free_y") +
    theme_void() +
    theme(legend.position = "none",
          text = element_text(face = "plain", family = regular),
          strip.text = element_text(face = "bold", family = regular))

}

if (build_figures) {
  ggsave(here("output", "figures", "figure_2_4.pdf"), 
         plot = figure_2_4_fun("Futura", "Futura Condensed"), 
         width = 8, height = 5, units = "in", useDingbats = FALSE)
  
  extrafont::embed_fonts(here("output", "figures", "figure_2_4.pdf"))
}

```

We adjudicate between these possibilities by comparing the activity of listings which had FREH status in January or February 2020—the months before the pandemic arrived—and listings which did not have this status. There were `r jan_feb_FREH_total` listings which we consider likely to have been FREH in either or both of January and February 2020. Of these listings, `r jan_feb_FREH_deactivated` were no longer listed on Airbnb or VRBO as of December 31, 2020. This is `r jan_feb_FREH_deactivated_pct` of these listings—almost twice as high as the `r jan_feb_FREH_deactivated_pct_2019` of listings which were FREH in either January or February 2019 and were no longer listed on the STR platforms by the end of August 2019. In total, `r jan_feb_non_FREH_deactivated_pct` of non-FREH listings active in January or February 2020 were deactivated by the end of December 2020, while the corresponding figure last year was `r jan_feb_non_FREH_deactivated_pct_2019`. This means that non-FREH listings have been deactivated at a higher rate this year than last year, but FREH listings have been deactivated at a substantially higher rate this year than last year (Figure \@ref(fig:fig-2-4)).

``` {r fig-2-4, include = TRUE, fig.cap = '(ref:fig-2-4)', fig.align = "center"}

figure_2_4_fun()

```

(ref:fig-2-4) _Activity trajectories of FREH and non-FREH listings in 2019 and 2020_

Of the `r jan_feb_FREH_active` FREH listings which remained listed throughout March - December, `r jan_feb_FREH_blocked_all` (`r jan_feb_FREH_blocked_all_pct`) were blocked (i.e. not available for reservations) for all of the month of December, and `r jan_feb_FREH_blocked_maj` (`r jan_feb_FREH_blocked_maj_pct`) were blocked for a majority of the month. This is extremely rare behaviour for a dedicated STR listing, since it is usually one of the busiest months for tourist accommodations in Toronto. In 2019, only `r jan_feb_FREH_blocked_all_pct_2019` of listings which were FREH in January or February were blocked for all of December, and only `r jan_feb_FREH_blocked_maj_pct_2019` were blocked for a majority of the month.

``` {r make_fig_2_5}

library(tsibble)

figure_2_5_fun <- function(regular = "", condensed = "") {
  
  fig_left <- 
    daily_reservation_trajectories %>% 
    group_by(FREH_feb) %>% 
    mutate(n = slider::slide_dbl(n, mean, .before = 13)) %>% 
    ungroup() %>% 
    ggplot(aes(date, n, colour = FREH_feb)) +
    geom_line(lwd = 1) +
    scale_x_date(name = NULL) +
    scale_y_continuous(name = "Total daily reservations", limits = c(0, NA), 
                       label = scales::comma) +
    scale_color_manual(name = "FREH status in January-February 2020", 
                       values = col_palette[c(5, 1)]) +
    theme_minimal() +
    theme(legend.position = "bottom", 
          panel.grid.minor.x = element_blank(),
          text = element_text(face = "plain", family = regular),
          legend.title = element_text(face = "bold", family = regular, 
                                      size = 10),
          legend.text = element_text( size = 10, family = regular))
  
  fig_right <- 
    monthly_reservation_trajectories %>% 
    mutate(date = as.Date(yearmon)) %>%
    ggplot(aes(date, n, colour = FREH_feb)) +
    geom_line(lwd = 1) +
    scale_x_date(name = NULL) +
    scale_y_continuous(name = "Average monthly reservations", limits = c(5, NA), 
                       label = scales::label_number(accuracy = 1)) +
    scale_color_manual(name = "FREH status in January-February 2020", 
                       values = col_palette[c(5, 1)]) +
    theme_minimal() +
    theme(legend.position = "bottom", 
          panel.grid.minor.x = element_blank(),
          text = element_text(face = "plain", family = regular),
          legend.title = element_text(face = "bold", family = regular,
                                      size = 10),
          legend.text = element_text(size = 10, family = regular))
  
  fig_left + fig_right + plot_layout(guides = 'collect') & 
    theme(legend.position = "bottom")

}

if (build_figures) {
  ggsave(here("output", "figures", "figure_2_5.pdf"), 
         plot = figure_2_5_fun("Futura", "Futura Condensed"), 
         width = 8, height = 5, units = "in", useDingbats = FALSE)
  
  extrafont::embed_fonts(here("output", "figures", "figure_2_5.pdf"))
}

```

Figure \@ref(fig:fig-2-5) compares the activity of the FREH and non-FREH listings which have remained active during the pandemic. The left panel shows the total reserved nights which occurred in each of these two groups. Unsurprisingly, it demonstrates that the large majority of reservations in the months prior to the pandemic occurred in these FREH properties. For example, in the month of February 2020, `r feb_2020_FREH_pct` of all reserved nights were booked in these FREH properties. In March 2020, reservations declined in both the FREH and non-FREH properties, but the decline is far larger in both relative and absolute terms among the FREH properties. The right panel aggregates the same data in a different fashion, showing the average number of booked nights per listing per month across the FREH and non-FREH listings. The same general pattern is observable: FREH listings receive substantially more bookings per month than non-FREH properties until March 2020, at which point both groups see their reservations decline, but the FREH listings decline much more sharply.

``` {r fig-2-5, include = TRUE, fig.cap = '(ref:fig-2-5)', fig.align = "center"}

figure_2_5_fun()

```

(ref:fig-2-5) _The total number of reservations per day (L) and the average reservations per listing per month (R), for listings which had FREH status in January or February 2020 and listings which did not (14-day average)_

The conclusion is that, while the entire STR market in Toronto has suffered an unprecedented collapse during the Covid-19 pandemic, this collapse has been disproportionately concentrated among dedicated STRs which have been responsible for thousands of units of rental housing lost in the city over the past several years. This raises the possibility that some of these units many have returned to the long-term rental market—a possibility that has been noted anecdotally in cities across Canada (McSheffrey 2020). We explore this possibility systematically in the next chapter.

## The impact of STRs on rental housing supply and vacancy rates

```{r para_4, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

qload(here("output", "str_processed.qsm"), nthreads = availableCores())
qload(here("output", "cmhc.qsm"), nthreads = availableCores())

# 2019 city vacancy
city_vacancy_2019 <- 
  city_vacancy %>% 
  filter(bedroom == "Total", date == 2019) %>% 
  pull(vacancy) %>% 
  scales::percent(0.1)

# Highest zone vacancy
highest_vacancy_zone <- 
  annual_vacancy %>% 
  filter(date == 2019, dwelling_type == "Total", bedroom == "Total", 
         quality == "a") %>% 
  filter(vacancy == max(vacancy)) %>% 
  pull(zone_name)

# Highest zone vacancy rate
highest_vacancy_zone_rate <- 
  annual_vacancy %>% 
  filter(date == 2019, dwelling_type == "Total", bedroom == "Total", 
         quality == "a") %>% 
  filter(vacancy == max(vacancy)) %>% 
  pull(vacancy) %>% 
  scales::percent(0.1)

# Highest zone vacancy rental units
highest_vacancy_zone_apts <- 
  annual_units %>% 
  filter(zone_name == highest_vacancy_zone, 
         date == 2019, dwelling_type == "Total", 
         bedroom == "Total") %>% 
  pull(units) %>%
  round(-2) %>% 
  prettyNum(",")

# Highest zone vacancy vacant rental units
highest_vacancy_zone_vac_apts <- 
  annual_units %>% 
  filter(zone_name == highest_vacancy_zone, 
         date == 2019, dwelling_type == "Total", 
         bedroom == "Total") %>% 
  pull(units) %>%
  {. * parse_number(highest_vacancy_zone_rate) / 100} %>% 
  round(-2) %>% 
  prettyNum(",")

# 3+ bedroom vacancy rate Toronto central
toronto_central_vac_3_bed <- 
  annual_vacancy %>% 
  filter(zone_name == "Toronto (Central)", date == 2019, dwelling_type == "Total",
         bedroom == "3 Bedroom +") %>% 
  pull(vacancy) %>% 
  scales::percent(0.1)

# Toronto West renter %
toronto_west_renters_pct <- 
  cmhc %>% 
  as_tibble() %>% 
  select(-geometry) %>% 
  mutate(p_renter = renter_households / total_households) %>% 
  select(zone_name, p_renter) %>% 
  filter(zone_name == "Toronto (West)") %>% 
  pull(p_renter) %>% 
  scales::percent(0.1)

# Toronto West owner %
toronto_west_owners_pct <- 
  cmhc %>% 
  as_tibble() %>% 
  select(-geometry) %>% 
  mutate(p_renter = renter_households / total_households) %>% 
  select(zone_name, p_renter) %>% 
  filter(zone_name == "Toronto (West)") %>% 
  pull(p_renter) %>% 
  {1 - .} %>% 
  scales::percent(0.1)

# Calculations for figure
renter_zone <- 
  cmhc %>% 
  as_tibble() %>% 
  select(-geometry) %>% 
  mutate(p_renter = renter_households / total_households) %>% 
  select(zone, p_renter)

daily_cmhc <- 
  property %>% 
  st_intersection(cmhc) %>% 
  st_drop_geometry() %>% 
  select(property_ID, zone) %>% 
  left_join(daily, .) %>% 
  filter(housing, date %in% as.Date(c("2019-12-01", "2019-12-31", "2018-12-01",
                                      "2018-12-31"))) %>% 
  mutate(FREH_3 = if_else(substr(date, 9, 9) == 0, FREH_3, 0),
         GH     = if_else(substr(date, 9, 9) == 3, GH, FALSE)) %>% 
  mutate(date = as.integer(substr(date, 1, 4))) %>% 
  group_by(zone, date) %>% 
  summarize(housing_loss = sum(FREH_3) + sum(GH)) %>% 
  ungroup()

strs_by_zone <- 
  property %>% 
  st_intersection(cmhc) %>% 
  st_drop_geometry() %>% 
  select(property_ID, zone) %>% 
  left_join(daily, .) %>% 
  filter(housing, date >= start_2019, date <= end_2019, 
         status != "B") %>% 
  count(zone) %>% 
  mutate(active_strs = n / 365)

unit_change <-
  annual_units %>% 
  filter(dwelling_type == "Total", bedroom == "Total") %>% 
  arrange(desc(date)) %>% 
  distinct(date, zone_name, .keep_all = TRUE) %>% 
  inner_join(daily_cmhc) %>% 
  left_join(renter_zone) %>%
  mutate(housing_loss = housing_loss * p_renter) %>%
  group_by(zone) %>% 
  summarize(unit_change = units[date == 2019] - units[date == 2018],
            housing_loss_change = housing_loss[date == 2019] - 
              housing_loss[date == 2018]) %>% 
  mutate(net_unit_change = unit_change - housing_loss_change) %>% 
  left_join(strs_by_zone) %>% 
  arrange(-active_strs) %>% 
  slice(1:10) %>% 
  left_join(st_drop_geometry(cmhc)) %>% 
  select(zone, zone_name, unit_change:net_unit_change)

total_unit_change <- 
  sum(unit_change$unit_change) %>% 
  round(-1)

# Vacancy calculations for map
vacancy_for_map <-
  annual_vacancy %>% 
  filter(dwelling_type == "Total", bedroom == "Total", !is.na(vacancy),
         quality == "a") %>% 
  group_by(zone) %>% 
  filter(date == max(date)) %>% 
  ungroup() %>% 
  left_join(cmhc) %>% 
  mutate(vacant_units = vacancy * renter_households) %>% 
  left_join(select(filter(daily_cmhc, date == 2019), zone, housing_loss)) %>% 
  left_join(renter_zone) %>% 
  filter(housing_loss >= 50) %>% 
  arrange(zone) %>% 
  mutate(units_returning = housing_loss * p_renter,
         new_vacant = vacant_units + units_returning,
         new_vacancy = new_vacant / renter_households) %>% 
  select(zone, zone_name, vacancy, new_vacancy, geometry) %>% 
  rename(`Current vacancy rate` = vacancy, 
         `Dedicated STRs back to market` = new_vacancy) %>% 
  pivot_longer(-c(zone, zone_name, geometry), names_to = "status",
               values_to = "vacancy") %>% 
  st_as_sf()


rm(property, daily, GH, annual_avg_rent, annual_units, annual_vacancy, 
   city_avg_rent, city_units, city_vacancy, cmhc)

```

In 2019, the City of Toronto’s rental vacancy rate stood at only `r city_vacancy_2019`—a crisis-level rate, albeit slightly higher than the previous few years. The standard rule of thumb is that a healthy rental market should have a vacancy rate of at least 3%, and all 17 of the CMHC's zones in Toronto are substantially below that. `r highest_vacancy_zone` has the highest vacancy rate at only `r highest_vacancy_zone_rate`, which means that, of the zone’s approximately `r highest_vacancy_zone_apts` rental apartments, fewer than `r highest_vacancy_zone_vac_apts` were available to be rented by prospective tenants in October 2019, when CMHC’s survey was conducted. In general, vacancy rates are even lower for family-sized housing units (defined by CMHC as units with two or more bedrooms). For example, the vacancy rate for units with three or more bedrooms was `r toronto_central_vac_3_bed` for the Toronto Central zone.

Given the thousands of housing units which we have documented as having been converted to dedicated STRs in Toronto, it is highly likely that the STR market is negatively affecting rental housing supply and the vacancy rate. One way to evaluate this possibility is to compare the net change in rental housing units in an area with the effective reductions in rental housing supply caused by the growth in dedicated STRs. Figure \@ref(fig:fig-2-6) makes this comparison, showing the change in primary rental housing units (i.e. apartments and townhouses, but not condominiums being used as rentals) per CMHC zone between 2018 and 2019, and the effective change in housing units once the impact of STRs taking housing off of the market is factored in. We assume that STR conversions happen in the same proportion as existing tenure arrangements in the neighbourhood. For example, in the Toronto West zone, `r toronto_west_renters_pct` of households are renters, so we assume that `r toronto_west_renters_pct` of housing units converted to dedicated STRs would have been rental housing, and the remaining `r toronto_west_owners_pct` would have been ownership housing. (We also assume CMHC has not removed dedicated STRs from their counts of primary rental units. In principle CMHC would not consider dedicated STRs to be rental units, but the difficulty in identifying dedicated STRs implies that the safest approach is to assume that dedicated STRs remain in the CMHC rental count.)

``` {r make_fig_2_6}

figure_2_6_fun <- function(regular = "", condensed = "") {
  
  unit_change %>%
    mutate(zone_name = factor(zone_name, levels = zone_name)) %>% 
    select(-zone) %>% 
    pivot_longer(-zone_name) %>% 
    filter(name != "housing_loss_change") %>%
    mutate(name = factor(name, 
                         levels = c("unit_change", "net_unit_change"))) %>% 
    ggplot() +
    geom_col(aes(zone_name, value, fill = name),
             position = position_dodge(width = 0.5)) +
    geom_segment(data = unit_change, 
                 mapping = aes(x = zone_name, xend = zone_name, y = unit_change, 
                               yend = net_unit_change),
                 size = 1.2,
                 arrow = arrow(length = unit(0.3, "cm")),
                 position = position_nudge(x = -0.125)) +
    scale_fill_manual(name = NULL, values = col_palette[c(3, 1)], labels = c(
      "Rental unit change", "Rental unit change with STRs")) +
    scale_y_continuous(name = NULL, breaks = c(-200, 0, 200, 400, 600)) +
    scale_x_discrete(name = NULL,
                     labels = unit_change$zone_name %>% 
                       str_replace_all("/", "/\n") %>% 
                       str_replace("east V", "east\nV") %>% 
                       str_replace("st H", "st\nH")) +
    theme_minimal() +
    theme(legend.position = "bottom",
          text = element_text(family = regular),
          axis.text.x = element_text(family = regular, size = 7),
          panel.grid.minor.x = element_blank(),
          panel.grid.major.x = element_blank())

}

if (build_figures) {
  ggsave(here("output", "figures", "figure_2_6.pdf"), 
         plot = figure_2_6_fun("Futura", "Futura Condensed"), 
         width = 8, height = 5, units = "in", useDingbats = FALSE)
  
  extrafont::embed_fonts(here("output", "figures", "figure_2_6.pdf"))
}

```

Between 2018 and 2019, the City of Toronto as a whole saw an `r total_unit_change`-unit increase, more than half of which occurred in Toronto Central. In a zone like Toronto Central, where there is a high concentration of STR operations and a growth in housing loss, the zone’s net rental housing change is greatly affected by new STR conversions in 2019. In almost every single of the ten selected CMHC zones (shown in Figure \@ref(fig:fig-2-6)), new STR conversions in 2019 either mostly cancelled out new rental housing production, or substantially decreased the net rental housing supply from 2018 to 2019.

``` {r fig-2-6, include = TRUE, fig.cap = '(ref:fig-2-6)', fig.align = "center"}

figure_2_6_fun()

```

(ref:fig-2-6) _Nominal and effective changes in rental housing supply, 2018-2019, by CMHC zone_

``` {r make_fig_2_7}

figure_2_7_fun <- function(regular = "", condensed = "") {
  
  vacancy_for_map %>% 
  mutate(label = scales::percent(vacancy, accuracy = 0.1)) %>% 
  ggplot() +
  geom_sf(data = province, colour = "transparent", fill = "grey93") +
  # geom_sf(data = streets, size = 0.2, colour = "white") +
  geom_sf(aes(fill = vacancy), colour = "white", alpha = 0.7) +
  geom_sf_label(aes(label = label), size = 2, family = regular) +
  scale_fill_gradientn(colors = col_palette[c(1, 3, 2)], 
                       na.value = "grey80",
                       labels = scales::percent)  +
  guides(fill = guide_colourbar(title = "% housing\nlost to STR",
                                title.vjust = 1)) + 
  facet_wrap(vars(status), nrow = 1) +
  gg_bbox(vacancy_for_map) +
  theme_void() +
  theme(legend.position = "none",
        text = element_text(family = regular, face = "plain"),
        legend.title = element_text(family = regular, face = "bold",
                                    size = 7),
        legend.title.align = 0.9,
        legend.text = element_text(family = regular, size = 5),
        strip.text = element_text(family = regular, face = "bold", size = 12),
        # panel.border = element_rect(colour = "white", size = 2)
        )

}

if (build_figures) {
  ggsave(here("output", "figures", "figure_2_7.pdf"), 
         plot = figure_2_7_fun("Futura", "Futura Condensed"), 
         width = 8, height = 4.2, units = "in", useDingbats = FALSE)
  
  extrafont::embed_fonts(here("output", "figures", "figure_2_7.pdf"))
}

```

Another way to contextualize the impact of housing units being converted to dedicated STRs is to examine the relationship between the quantity of dedicated STRs and the rental vacancy rate. The left panel of Figure \@ref(fig:fig-2-7) displays the 2019 rental vacancy rate in the 10 CMHC zones with the largest number of active STR listings. The right panel displays a hypothetical vacancy rate if all dedicated STRs returned to the long-term market in these neighbourhoods. As with the calculations of net housing supply changes above, we assume that these units return to the market in the same proportion as existing tenure arrangements in the neighbourhood. The map shows that, in virtually every zone, the number of dedicated STRs is enough to dramatically increase housing supply if they were to be returned to the housing market. In every zone, the vacancy rate could double. 

``` {r fig-2-7, include = TRUE, fig.cap = '(ref:fig-2-7)', fig.align = "center"}

figure_2_7_fun()

```

(ref:fig-2-7) _2019 vacancy rate in the City of Toronto (L), and hypothetical vacancy rate if 2019 dedicated STRs returned to rental market (R), by CMHC zone_

To be clear, in a scenario where thousands of dedicated STRs returned to the long-term market, these units would not actually remain vacant, but would be absorbed into the rental housing supply. Eventually the result would be a combination of lower rents and a vacancy rate lower than depicted in the map but higher than the current vacancy rate. Nevertheless, Figure \@ref(fig:fig-2-7) demonstrates that the number of housing units which have left Toronto’s long-term market to be operated as STRs is sufficient to significantly constrain housing availability in the city, in 2019 levels.

\newpage