# Bringing back additional STR onto the LTR market

<style type="text/css">
  body{
  font-family: Futura, Helvetica, Arial;
}
</style>

<br>

```{r setup, include = FALSE, echo = FALSE}

###### for max to be able to knit
if(memory.limit()< 20000) {memory.limit(size = 48000)}
####

library(here)
source(here("R", "01_startup.R"))
library(imager)

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(include = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# Function for checking file modifications
mtime <- function(files) lapply(Sys.glob(files), function(x) file.info(x)$mtime)

# Set this to TRUE to output PDF figures
build_figures <- FALSE

```

```{r new_objects, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

qload("output/str_processed.qsm", nthreads = availableCores())
ltr <- qread("output/ltr_processed.qs", nthreads = availableCores())

ltr_unique_price <- 
  ltr %>% 
  st_drop_geometry() %>% 
  filter(!is.na(price)) %>% 
  # keep the more recent price
  arrange(desc(scraped)) %>% 
  distinct(id, .keep_all = T) %>% 
  select(ltr_ID = id, ltr_price = price, ltr_scraped = scraped, ward, property_ID)

# Monthly revenue in 2019
revenue_2019 <-
  daily %>% 
  filter(date >= start_2019, date <= end_2019,
         status == c("A", "R")) %>% 
  mutate(month = month(date)) %>%
  group_by(property_ID, month) %>% 
  filter(n() >= 5) %>% 
  filter(status == "R") %>% 
  summarize(monthly_revenue_2019 = sum(price)) %>%
  group_by(property_ID) %>%
  summarize(monthly_revenue_2019 = mean(monthly_revenue_2019)) %>% 
  ungroup()

# Monthly revenue in 2020
revenue_2020 <-
  daily %>% 
  filter(date >= start_2020, date <= end_2020,
         status == c("A", "R")) %>% 
  mutate(month = month(date)) %>%
  group_by(property_ID, month) %>% 
  filter(n() >= 5) %>% 
  filter(status == "R") %>% 
  summarize(monthly_revenue_2020 = sum(price)) %>%
  group_by(property_ID) %>%
  summarize(monthly_revenue_2020 = mean(monthly_revenue_2020)) %>% 
  ungroup() %>% 
  # was the listing found on the LTR?
  left_join(select(st_drop_geometry(property), property_ID, ltr_ID)) %>% 
  mutate(ltr = ifelse(!is.na(ltr_ID), T, F)) %>% 
  select(-ltr_ID) 

# # Monthly revenue variation between 2019 and 2020
monthly_revenue_variation <-
revenue_2020 %>%
  left_join(revenue_2019) %>%
  mutate(monthly_variation = (monthly_revenue_2020-monthly_revenue_2019)/monthly_revenue_2019)

# Property status (still on the platform or not?)
property_status <- 
  property %>% 
  st_drop_geometry() %>% 
  filter(!is.na(ltr_ID)) %>%
  filter(scraped >= "2020-01-01") %>% 
  mutate(status = case_when(scraped <= max(scraped)-days(30) ~ "Taken down", # TKTK date maybe to update, depends on the last scraped date
                            active < "2020-12-01" ~ "Blocked", # Not active during all october
                            TRUE ~ "Continued")) # what if active = NA ?
  

# Matrice creation
matrice <-
  monthly_revenue_variation %>%
  left_join(unnest(select(property_status, property_ID, ltr_ID, status, bedrooms, ward), ltr_ID),
             by = "property_ID") %>% 
  left_join(select(ltr_unique_price, -property_ID, -ward), by = "ltr_ID") %>% 
  # Keep the most recent LTR price scraped per property_ID
  arrange(desc(ltr_scraped)) %>% 
  distinct(property_ID, .keep_all=T) %>% 
  select(-ltr_scraped) %>% 
  mutate(multi = ifelse(property_ID %in% (daily %>% filter(multi) %>% pull(property_ID) %>% unique()), T, F),
         FREH = ifelse(property_ID %in% (daily %>% filter(FREH_3>0.5) %>% pull(property_ID) %>% unique()), T, F),
         commercial = ifelse(multi== T | FREH == T, T, F))

rm(daily, GH, ltr, monthly_revenue_variation, property, 
   property_status, revenue_2019, revenue_2020)

```

**Using a multivariate linear regression model, we estimated specific reasonable amounts at which a STR listing's host would accept to move his unit to the LTR market, using data from web scrapes and the previously explained image recognition techniques, used to build a training dataset. Our significant variables used are the 2020 monthly revenue, the size and the location of the unit.**

## Relative financial performance of STR moving to the LTR market

As shown in the previous section, the COVID-19 pandemic and the expectations regarding the new regulation have resulted in hundreds of hosts listing their short-term rental units on the long-term rental market. This final chapter aims at understanding the financial motivations for why a host would choose to operate this change, and to establish a relationship between financial performance, on the STR platforms, and asking rents.

```{r monthly_rev, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

monthly_rev_median_ltr <- 
matrice %>% 
    group_by(ltr) %>% 
    summarize(monthly_revenue_2020_median = median(monthly_revenue_2020, na.rm = T),
              monthly_revenue_2019_median = median(monthly_revenue_2019, na.rm = T),
              monthly_variation_median = median(monthly_variation, na.rm = T)) %>% 
  mutate(across(where(is.numeric), round, 2))

```

``` {r make_fig_5_1}

figure_5_1_fun <- function(regular = "", condensed = "") {

matrice %>% 
  ggplot()+
  geom_histogram(aes(monthly_revenue_2020, y=..density.., fill = ltr), position = "dodge") +
  geom_vline(data = monthly_rev_median_ltr,
             aes(xintercept = monthly_revenue_2020_median, color = ltr),
             linetype="dashed", size = 1, show.legend = FALSE)+
  xlim(0,3000)+
  scale_fill_manual(name = NULL, labels = c("Not matched on the LTR",
                                             "Matched on the LTR"),
                     values= col_palette[c(4,2)]) +
  ylab("Density")+
  xlab("Monthly revenue in 2020")+
  theme_minimal() +
    theme(legend.position = "bottom",
          panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          text = element_text(face = "plain", family = regular),
          legend.title = element_text(face = "bold", family = regular,
                                      size = 10),
          legend.text = element_text(size = 10, family = regular),
          strip.text = element_text(face = "bold", family = regular))

}

if (build_figures) {
  ggsave(here("output", "figures", "figure_5_1.pdf"),
         plot = figure_6_1_fun("Futura", "Futura Condensed"),
         width = 8, height = 5, units = "in", useDingbats = FALSE)

  extrafont::embed_fonts(here("output", "figures", "figure_5_1.pdf"))
}

```

``` {r fig-5-1, include = TRUE, fig.cap = '(ref:fig-5-1)', fig.align = "center"}

figure_5_1_fun()

```

(ref:fig-5-1) _Monthly revenue in 2020 for matched and not matched properties_

STR listings placed on the LTR market had significant higher monthly revenues in 2020 compared to the STR listings that did not make the move to the LTR market. Indeed, while the non-matched listings had a median revenue (the dashed lines on Figure \@ref(fig:fig-5-1)) of \$`r monthly_rev_median_ltr[1,2]`, the listings that did match have made $`r monthly_rev_median_ltr[2,2]`. The difference is even greater when looking at the median monthly revenue of the precedent year (2019) for the same listings, when they respectively made \$`r monthly_rev_median_ltr[1,3]` and \$`r monthly_rev_median_ltr[2,3]`. The drop in revenue between the two years, for both group, is also notable. Figure \@ref(fig:fig-5-1) demonstrates how the STR listings that did not match a LTR listing had significant lower monthly revenues in 2020. It is in line with our previous assessment, stating that it is the most affluent hosts which chose to move their lucrative units on the LTR market.

``` {r make_fig_5_2}

figure_5_2_fun <- function(regular = "", condensed = "") {

matrice %>% 
  ggplot()+
  geom_density(aes(monthly_variation, color = ltr), size = 1) +
  geom_vline(data = monthly_rev_median_ltr,
             aes(xintercept = monthly_variation_median, color = ltr),
             linetype="dashed", size = 1)+
  xlim(-1,1.2)+
  scale_color_manual(name = NULL, labels = c("Not matched on the LTR", 
                                             "Matched on the LTR"),
                     values= col_palette[c(4,2)]) +
  ylab("Density")+
  xlab("Monthly revenue variation between 2019 and 2020")+
  theme_minimal() +
    theme(legend.position = "bottom",
          panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          text = element_text(face = "plain", family = regular),
          legend.title = element_text(face = "bold", family = regular,
                                      size = 10),
          legend.text = element_text(size = 10, family = regular),
          strip.text = element_text(face = "bold", family = regular))

}

if (build_figures) {
  ggsave(here("output", "figures", "figure_5_2.pdf"),
         plot = figure_6_1_fun("Futura", "Futura Condensed"),
         width = 8, height = 5, units = "in", useDingbats = FALSE)

  extrafont::embed_fonts(here("output", "figures", "figure_5_2.pdf"))
}

```

``` {r fig-5-2, include = TRUE, fig.cap = '(ref:fig-5-2)', fig.align = "center"}

figure_5_2_fun()

```

(ref:fig-5-2) _Variation of monthly revenue between 2019 and 2020 for matched and not matched properties_

Monthly revenues in 2020 is not the only relevant factor to consider when looking to explain the behavior of a host moving its STR unit to the LTR market. Indeed, it may be useful to see the relative loss of monthly revenues between 2019 and 2020, again separated between the two groups previously evaluated. While the vast majority of listings had a decrease in monthly revenue, it is the listings matched on the LTR that have had a worse relative loss in monthly revenue, as Figure \@ref(fig:fig-5-2) shows. In brief, listings found on the LTR market made a higher monthly revenue in 2020 than the other listings, but had a greater decrease in monthly revenue when compared to 2019's monthly revenue levels.

## Breakdown into groups and financial analysis of the matched STR listings

```{r filter_outliers, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

matrice <- 
  matrice %>% 
  filter(ltr_price < as.numeric(quantile(matrice$ltr_price, 0.975, na.rm = T)),
         ltr_price > as.numeric(quantile(matrice$ltr_price, 0.025, na.rm = T)),
         monthly_revenue_2020 < as.numeric(quantile(matrice$monthly_revenue_2020,
                                                    0.975, na.rm = T)),
         monthly_revenue_2020 > as.numeric(quantile(matrice$monthly_revenue_2020,
                                                    0.025, na.rm = T)))

monthly_rev_median_status <- 
  matrice %>% 
  group_by(status) %>% 
  summarize(n = n(),
            monthly_revenue_2019_median = median(monthly_revenue_2019, na.rm = T),
            monthly_revenue_2020_median = median(monthly_revenue_2020, na.rm = T),
            monthly_variation_median = median(monthly_variation, na.rm = T)) %>% 
  arrange(monthly_revenue_2020_median) %>% 
  mutate(across(where(is.numeric), round, 2))

```

As in the precedent chapter, we separated here the matched listings into three groups: The blocked listings, the ones that were taken down, and the ones that continued their operations on the STR market. This analysis is based on properties listed on the LTR market, and having at least a month of activity in 2020 (a total of `r sum(monthly_rev_median_status$n)` listings).

``` {r make_fig_5_3}

figure_5_3_fun <- function(regular = "", condensed = "") {

matrice %>% 
  ggplot()+
  geom_density(aes(monthly_revenue_2020, color = status), size = 1)+
  geom_vline(data = monthly_rev_median_status,
             aes(xintercept = monthly_revenue_2020_median, color = status),
             linetype="dashed", size = 1)+
  # facet_wrap(~status)+
  scale_color_manual(name = NULL, 
                     values= col_palette[c(6,4,2)])+
  ylab("Density")+
  xlab("Matched listings' monthly revenue in 2020")+
  theme_minimal() +
    theme(legend.position = "bottom",
          panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          text = element_text(face = "plain", family = regular),
          legend.title = element_text(face = "bold", family = regular,
                                      size = 10),
          legend.text = element_text(size = 10, family = regular),
          strip.text = element_text(face = "bold", family = regular))

}

if (build_figures) {
  ggsave(here("output", "figures", "figure_5_3.pdf"),
         plot = figure_6_1_fun("Futura", "Futura Condensed"),
         width = 8, height = 5, units = "in", useDingbats = FALSE)

  extrafont::embed_fonts(here("output", "figures", "figure_5_3.pdf"))
}

```

``` {r fig-5-3, include = TRUE, fig.cap = '(ref:fig-5-3)', fig.align = "center"}

figure_5_3_fun()

```

(ref:fig-5-3) _Matched listings' 2020 monthly revenue, separated in three groups_

The listings that matched but continued their STR operations are the listings with the smallest monthly revenue in 2020 (a median of \$`r monthly_rev_median_status[1,4]`), while the blocked and taken down listings earned more, with respectively \$`r monthly_rev_median_status[2,4]` and \$`r monthly_rev_median_status[3,4]` of median monthly revenue in 2020. The three numbers are significantly lower than what a landlord could make on the LTR market, which is a motivation to operate the change from STR to LTR. It is the listings that did not succeed their transfer that made a smaller monthly revenue in 2020, probably because they operated longer under the overall COVID situation and during fall, where revenue are overall decreasing on the platform. The listings that are completely taken down are the listings that made the most monthly revenue, TKTK.

## Modeling the relationship between listings' characteristics and asking rents

How can the city bring more dedicated STR housing units to the LTR market? We can use the population of STR listings matched to the LTR to infer what would be a plausible asking rent for the other STR listings, if they were to transfer to the LTR market. We found three significant variables that can be used to do so. Firstly, the financial performance in 2020. Indeed, there is a significant positive relationship between the 2020 monthly revenue and the asked rent, as seen in Figure \@ref(fig:fig-5-4). There is no notable discrepancy between the three categories (blocked, continued or taken down), so the dark blue regression line combines all three.

``` {r make_fig_5_4}

figure_5_4_fun <- function(regular = "", condensed = "") {

matrice %>% 
  ggplot()+
  geom_point(aes(monthly_revenue_2020, ltr_price, color = status), size =1)+
  geom_smooth(method = "lm",  aes(monthly_revenue_2020, ltr_price), 
              size = 1, se = F, color = col_palette[1])+
  scale_color_manual(name = NULL, 
                     values= col_palette[c(6,4,2)])+
  ylab("Asking rent")+
  xlab("Monthly revenue in 2020")+
  theme_minimal() +
    theme(legend.position = "bottom",
          panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          text = element_text(face = "plain", family = regular),
          legend.title = element_text(face = "bold", family = regular,
                                      size = 10),
          legend.text = element_text(size = 10, family = regular),
          strip.text = element_text(face = "bold", family = regular))

}

if (build_figures) {
  ggsave(here("output", "figures", "figure_5_4.pdf"),
         plot = figure_6_1_fun("Futura", "Futura Condensed"),
         width = 8, height = 5, units = "in", useDingbats = FALSE)

  extrafont::embed_fonts(here("output", "figures", "figure_5_4.pdf"))
}

```

``` {r fig-5-4, include = TRUE, fig.cap = '(ref:fig-5-4)', fig.align = "center"}

figure_5_4_fun()

```

(ref:fig-5-4) _Positive relationship between monthly revenue asked rent_

```{r ward_higher_rev, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

ward_price_bounds <- 
  ltr_unique_price %>% 
  filter(ltr_price < as.numeric(quantile(ltr_unique_price$ltr_price, 0.99)),
         ltr_price > as.numeric(quantile(ltr_unique_price$ltr_price, 0.025))) %>% 
  group_by(ward) %>% 
  summarize(ltr_mean_price = mean(ltr_price), n()) %>% 
  mutate(geo_quant = percent_rank(ltr_mean_price)) %>%
  mutate(geo_quant = case_when(geo_quant <= 0.75 ~ "lower",
                               # geo_quant <= 0.66 ~ "middle",
                               TRUE ~ "higher"),
         geo_quant = factor(geo_quant, levels = c("lower",# "middle", 
                                                  "higher")))
wards_higher_bound <- 
ward_price_bounds %>% 
  filter(geo_quant == "higher") %>% 
  arrange(-ltr_mean_price) %>% 
  pull(ward) %>%
  paste(collapse = ", ")

```

The next variable we found to be significant for our model is the size of the units. No surprise, the number of bedrooms announced in the STR  listing has an impact on the rent asked on the LTR platforms.

The last variable is location. Intrinsically, the monthly revenue variable holds a location characteristics: listings in more affluent locations, say the waterfront, will, on average, make more revenue. However, when analyzing all the scraped LTR listings, we found that the wards of `r wards_higher_bound`, were the wards with the highest asking rent on average. We also found that the matched listings from these wards were listed at a higher rent, independently of the size and the financial performance of their respective STR listings. So, we added this location variable to our model. 

```{r statistical_model, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

listings_pred <- read_csv("output/listings_pred.csv")

mean_pred_bedrooms <- 
listings_pred %>% 
  filter(`Number of bedrooms` <= 4) %>% 
  group_by(`Number of bedrooms`) %>% 
  summarize(`Mean asking rent predicted` = mean(`Predicted asking rent`))

rm(listings_pred)

```

The model offering an estimation of asking rent if a STR was to move to the LTR was applied to the entire population of entire home STR in Toronto and has been given to the City. For example, Table \@ref(tab:tab-pred) presents the mean asking rent predicted for each size of units. Our estimations of asking rent for all listings also depends on financial performance and location of each listings. 

``` {r tab-pred, include = TRUE}

library(kableExtra)

mean_pred_bedrooms %>% 
  mutate(across(where(is.numeric), round, 2)) %>%  
  kbl(caption = "Mean asking rent predicted by number of bedrooms", align = "lrrrr") %>% 
  kable_styling()

```


\newpage