# Bringing back additional STR onto the LTR market

<style type="text/css">
  body{
  font-family: Futura, Helvetica, Arial;
}
</style>

<br>

```{r setup, include = FALSE, echo = FALSE}

###### for max to be able to knit
if(memory.limit()< 20000) {memory.limit(size = 48000)}
####

library(here)
source(here("R", "01_startup.R"))
library(imager)

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(include = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# Function for checking file modifications
mtime <- function(files) lapply(Sys.glob(files), function(x) file.info(x)$mtime)

# Set this to TRUE to output PDF figures
build_figures <- TRUE

```

```{r new_objects, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

qload("output/str_processed.qsm", nthreads = availableCores())
ltr <- qread("output/ltr_processed.qs", nthreads = availableCores())

ltr_unique_price <- 
  ltr %>% 
  st_drop_geometry() %>% 
  filter(!is.na(price)) %>% 
  # keep the more recent price
  arrange(desc(scraped)) %>% 
  distinct(id, .keep_all = T) %>% 
  select(ltr_ID = id, ltr_price = price, ltr_scraped = scraped, ward, property_ID)

# Monthly revenue in 2019
revenue_2019 <-
  daily %>% 
  filter(date >= start_2019, date <= end_2019,
         status == c("A", "R")) %>% 
  mutate(month = month(date)) %>%
  group_by(property_ID, month) %>% 
  filter(n() >= 5) %>% 
  filter(status == "R") %>% 
  summarize(monthly_revenue_2019 = sum(price)) %>%
  group_by(property_ID) %>%
  summarize(monthly_revenue_2019 = mean(monthly_revenue_2019)) %>% 
  ungroup()

# Monthly revenue in 2020
revenue_2020 <-
  daily %>% 
  filter(date >= start_2020, date <= end_2020,
         status == c("A", "R")) %>% 
  mutate(month = month(date)) %>%
  group_by(property_ID, month) %>% 
  filter(n() >= 5) %>% 
  filter(status == "R") %>% 
  summarize(monthly_revenue_2020 = sum(price)) %>%
  group_by(property_ID) %>%
  summarize(monthly_revenue_2020 = mean(monthly_revenue_2020)) %>% 
  ungroup() %>% 
  # was the listing found on the LTR?
  left_join(select(st_drop_geometry(property), property_ID, ltr_ID)) %>% 
  mutate(ltr = ifelse(!is.na(ltr_ID), T, F)) %>% 
  select(-ltr_ID) 

# # Monthly revenue variation between 2019 and 2020
monthly_revenue_variation <-
revenue_2020 %>%
  left_join(revenue_2019) %>%
  mutate(monthly_variation = (monthly_revenue_2020 - monthly_revenue_2019) / 
           monthly_revenue_2019)

# Property status (still on the platform or not?)
property_status <- 
property %>% 
  st_drop_geometry() %>% 
  filter(!is.na(ltr_ID)) %>%
  filter(scraped >= "2020-01-01") %>% 
  mutate(status = case_when(scraped <= max(scraped)-days(30) ~ "Taken down", # TKTK date maybe to update, depends on the last scraped date
                            active < "2020-12-01" ~ "Blocked", # Not active during all october
                            TRUE ~ "Continued STR"),
         status = ifelse(status == "Continued STR" & minimum_stay >= 28, "28-days min.", status))
  

# Matrice creation
matrice <-
  monthly_revenue_variation %>%
  left_join(unnest(select(property_status, property_ID, ltr_ID, status, bedrooms, ward), ltr_ID),
             by = "property_ID") %>% 
  left_join(select(ltr_unique_price, -property_ID, -ward), by = "ltr_ID") %>% 
  # Keep the most recent LTR price scraped per property_ID
  arrange(desc(ltr_scraped)) %>% 
  distinct(property_ID, .keep_all=T) %>% 
  select(-ltr_scraped) %>% 
  mutate(multi = ifelse(property_ID %in% (daily %>% filter(multi) %>% pull(property_ID) %>% unique()), T, F),
         FREH = ifelse(property_ID %in% (daily %>% filter(FREH_3>0.5) %>% pull(property_ID) %>% unique()), T, F),
         commercial = ifelse(multi== T | FREH == T, T, F))

rm(daily, GH, ltr, monthly_revenue_variation, property, 
   property_status, revenue_2019, revenue_2020)

```

**Using a multivariate linear regression model, we estimated specific reasonable amounts at which a STR listing's host would accept to move his unit to the LTR market, using data from web scrapes and the previously explained image recognition techniques, used to build a training dataset. Our significant variables used are the 2020 monthly revenue, the size and the location of the unit.**

Chapter 3, above, established that some combination of the COVID-19 pandemic and the impending implementation of the City of Toronto's STR regulations have resulted in thousands of Toronto STR listing their STR units on the long-term rental market. This final chapter analyzes the financial parameters underlying this decision, and develops a quantitative model to predict the monthly rents at which the hosts of existing dedicated STRs might be induced to convert their listings to long-term rentals, based on the listing's previous financial performance, its size, and its location.

## The financial performance of STRs which moved to the LTR market

```{r monthly_rev, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

monthly_rev_median_ltr <- 
  matrice %>% 
  group_by(ltr) %>% 
  summarize(monthly_revenue_2020_median = median(monthly_revenue_2020, na.rm = T),
            monthly_revenue_2019_median = median(monthly_revenue_2019, na.rm = T),
            monthly_variation_median = median(monthly_variation, na.rm = T)) %>% 
  mutate(across(c(2:3), scales::dollar, 10))

```

``` {r make_fig_5_1}

figure_5_1_fun <- function(regular = "", condensed = "") {

top_plot <- 
  matrice %>% 
  ggplot() +
  geom_histogram(aes(monthly_revenue_2020, y=..density.., fill = ltr), 
                 position = position_dodge(width = 60), alpha = 0.8) +
  geom_vline(data = monthly_rev_median_ltr,
             aes(xintercept = monthly_revenue_2020_median, color = ltr),
             linetype="dashed", size = 1, show.legend = FALSE) +
  scale_fill_manual(name = NULL, labels = c("Not matched to LTR", "Matched to LTR"),
                    values= col_palette[c(4,2)]) +
  scale_x_continuous(name = "Monthly STR revenue (2020)", 
                     labels = scales::dollar, limits = c(0, 3000)) +
  scale_y_continuous(name = "Probability density", labels = NULL) +
  theme_minimal() +
    theme(legend.position = "bottom",
          panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          text = element_text(face = "plain", family = regular),
          legend.title = element_text(face = "bold", family = regular,
                                      size = 10),
          legend.text = element_text(size = 10, family = regular),
          strip.text = element_text(face = "bold", family = regular))

bottom_plot <- 
  matrice %>% 
  ggplot() +
  geom_histogram(aes(monthly_variation, y=..density.., fill = ltr), 
                 position = position_dodge(width = 60), alpha = 0.8) +
  geom_vline(data = mutate(monthly_rev_median_ltr, across(where(is.character), parse_number)),
             aes(xintercept = monthly_variation_median, color = ltr),
             linetype="dashed", size = 1, show.legend = FALSE) +
  xlim(-1,1.2)+
  scale_color_manual(name = NULL, labels = c("Not matched to LTR", "Matched to LTR"),
                     values = col_palette[c(4,2)]) +
  ylab("Density")+
  xlab("Monthly revenue variation between 2019 and 2020") +
  guides(colour = guide_legend(override.aes = list(fill = col_palette[c(4,2)]))) +
  theme_minimal() +
    theme(legend.position = "bottom",
          panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          text = element_text(face = "plain", family = regular),
          legend.title = element_text(face = "bold", family = regular,
                                      size = 10),
          legend.text = element_text(size = 10, family = regular),
          strip.text = element_text(face = "bold", family = regular))

top_plot + bottom_plot + plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")

}

if (build_figures) {
  ggsave(here("output", "figures", "figure_5_1.pdf"),
         plot = figure_5_1_fun("Futura", "Futura Condensed"),
         width = 8, height = 5, units = "in", useDingbats = FALSE)

  extrafont::embed_fonts(here("output", "figures", "figure_5_1.pdf"))
}

```

STR listings moved to the LTR market had significant higher monthly STR revenues in 2020 compared to the STR listings that did not make the move. Non-matched listings had a median revenue (the dashed lines in the upper panel of Figure \@ref(fig:fig-5-1)) of `r monthly_rev_median_ltr[1,2]`, while listings that did match to the LTR market had a median revenue of `r monthly_rev_median_ltr[2,2]`. The difference is larger for pre-pandemic 2019 monthly income, when non-matched and matched listings had median monthly incomes respectively of `r monthly_rev_median_ltr[1,3]` and `r monthly_rev_median_ltr[2,3]`. This pattern is consistent with our previous assessment that it is high-earning STR hosts who disproportionately chose to move their units to the LTR market. Moreover, while the vast majority of listings experienced a decrease in revenue from 2019 to 2020 because of the pandemic, it is the listings that moved to the long-term market which, on average, experienced larger declines, as the bottom panel in Figure \@ref(fig:fig-5-1) shows. In sum, relative to other listings, STR listings which transitioned to the LTR market had higher monthly revenue in 2020 but a greater decrease in year-over-year revenue.

``` {r fig-5-1, include = TRUE, fig.cap = '(ref:fig-5-1)', fig.align = "center"}

figure_5_1_fun()

```

(ref:fig-5-1) _Monthly revenue in 2020 (top panel), and variation of monthly revenue between 2019 and 2020 (bottom panel) for matched and not matched properties_


## Breakdown into groups and financial analysis of the matched STR listings

```{r filter_outliers, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

matrice <- 
  matrice %>% 
  filter(ltr_price < as.numeric(quantile(matrice$ltr_price, 0.975, na.rm = T)),
         ltr_price > as.numeric(quantile(matrice$ltr_price, 0.025, na.rm = T)),
         monthly_revenue_2020 < as.numeric(quantile(matrice$monthly_revenue_2020,
                                                    0.975, na.rm = T)),
         monthly_revenue_2020 > as.numeric(quantile(matrice$monthly_revenue_2020,
                                                    0.025, na.rm = T)))

monthly_rev_median_status <- 
  matrice %>% 
  group_by(status) %>% 
  summarize(n = n(),
            monthly_revenue_2019_median = median(monthly_revenue_2019, na.rm = T),
            monthly_revenue_2020_median = median(monthly_revenue_2020, na.rm = T),
            monthly_variation_median = median(monthly_variation, na.rm = T)) %>% 
  arrange(monthly_revenue_2020_median) %>% 
  mutate(across(where(is.numeric), round, 2))

```

As in the precedent chapter, we separated here the matched listings into three groups: The blocked listings, the ones that were taken down, and the ones that continued their operations on the STR market. This analysis is based on properties listed on the LTR market, and which were active at least 5 days during a month in 2020 (a total of `r sum(monthly_rev_median_status$n)` listings).

``` {r make_fig_5_3}

figure_5_3_fun <- function(regular = "", condensed = "") {

matrice %>% 
  ggplot()+
  geom_density(aes(monthly_revenue_2020, color = status), size = 1)+
  geom_vline(data = monthly_rev_median_status,
             aes(xintercept = monthly_revenue_2020_median, color = status),
             linetype="dashed", size = 1)+
  # facet_wrap(~status)+
  scale_color_manual(name = NULL, 
                     values= col_palette[c(6,4,2)])+
  ylab("Density")+
  xlab("Matched listings' monthly revenue in 2020")+
  theme_minimal() +
    theme(legend.position = "bottom",
          panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          text = element_text(face = "plain", family = regular),
          legend.title = element_text(face = "bold", family = regular,
                                      size = 10),
          legend.text = element_text(size = 10, family = regular),
          strip.text = element_text(face = "bold", family = regular))

}

if (build_figures) {
  ggsave(here("output", "figures", "figure_5_3.pdf"),
         plot = figure_5_3_fun("Futura", "Futura Condensed"),
         width = 8, height = 5, units = "in", useDingbats = FALSE)

  extrafont::embed_fonts(here("output", "figures", "figure_5_3.pdf"))
}

```

``` {r fig-5-3, include = TRUE, fig.cap = '(ref:fig-5-3)', fig.align = "center"}

figure_5_3_fun()

```

(ref:fig-5-3) _Matched listings' 2020 monthly revenue, separated in three groups_

The listings that matched but continued their operations with a minimum stay of 28 days are the listings with the smallest monthly revenue in 2020 (a median of \$`r monthly_rev_median_status[1,4]`), while the continued STR operations made a similar revenue (\$`r monthly_rev_median_status[2,4]`). Taken down and blocked listings earned more, with respectively \$`r monthly_rev_median_status[3,4]` and \$`r monthly_rev_median_status[4,4]` of median monthly revenue in 2020. The four numbers are significantly lower than what a landlord could make on the LTR market, which is a motivation to operate the change from STR to LTR. It is the listings that did not succeed their transfer that made a smaller monthly revenue in 2020, probably because they operated longer under the overall COVID situation and during fall, where revenue are overall decreasing on the platform. The listings that are completely taken down are the listings that made the most monthly revenue, TKTK.

## Modeling the relationship between listings' characteristics and asking rents

How can the city bring more dedicated STR housing units to the LTR market? We can use the population of STR listings matched to the LTR to infer what would be a plausible asking rent for the other STR listings, if they were to transfer to the LTR market. We found three significant variables that can be used to do so. Firstly, the financial performance in 2020. Indeed, there is a significant positive relationship between the 2020 monthly revenue and the asked rent, as seen in Figure \@ref(fig:fig-5-4). There is no notable discrepancy between the three categories (blocked, continued or taken down), so the dark blue regression line combines all three.

``` {r make_fig_5_4}

figure_5_4_fun <- function(regular = "", condensed = "") {

matrice %>% 
  ggplot()+
  geom_point(aes(monthly_revenue_2020, ltr_price, color = status), size =1)+
  geom_smooth(method = "lm",  aes(monthly_revenue_2020, ltr_price), 
              size = 1, se = F, color = col_palette[1])+
  scale_color_manual(name = NULL, 
                     values= col_palette[c(6,4,2)])+
  ylab("Asking rent")+
  xlab("Monthly revenue in 2020")+
  theme_minimal() +
    theme(legend.position = "bottom",
          panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          text = element_text(face = "plain", family = regular),
          legend.title = element_text(face = "bold", family = regular,
                                      size = 10),
          legend.text = element_text(size = 10, family = regular),
          strip.text = element_text(face = "bold", family = regular))

}

if (build_figures) {
  ggsave(here("output", "figures", "figure_5_4.pdf"),
         plot = figure_5_4_fun("Futura", "Futura Condensed"),
         width = 8, height = 5, units = "in", useDingbats = FALSE)

  extrafont::embed_fonts(here("output", "figures", "figure_5_4.pdf"))
}

```

``` {r fig-5-4, include = TRUE, fig.cap = '(ref:fig-5-4)', fig.align = "center"}

figure_5_4_fun()

```

(ref:fig-5-4) _Positive relationship between monthly revenue asked rent_

```{r ward_higher_rev, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

ward_price_bounds <- 
  ltr_unique_price %>% 
  filter(ltr_price < as.numeric(quantile(ltr_unique_price$ltr_price, 0.99)),
         ltr_price > as.numeric(quantile(ltr_unique_price$ltr_price, 0.025))) %>% 
  group_by(ward) %>% 
  summarize(ltr_mean_price = mean(ltr_price), n()) %>% 
  mutate(geo_quant = percent_rank(ltr_mean_price)) %>%
  mutate(geo_quant = case_when(geo_quant <= 0.75 ~ "lower",
                               # geo_quant <= 0.66 ~ "middle",
                               TRUE ~ "higher"),
         geo_quant = factor(geo_quant, levels = c("lower",# "middle", 
                                                  "higher")))
wards_higher_bound <- 
ward_price_bounds %>% 
  filter(geo_quant == "higher") %>% 
  arrange(-ltr_mean_price) %>% 
  pull(ward) %>%
  paste(collapse = ", ")

```

The next variable we found to be significant for our model is the size of the units. No surprise, the number of bedrooms announced in the STR  listing has an impact on the rent asked on the LTR platforms.

The last variable is location. Intrinsically, the monthly revenue variable holds a location characteristics: listings in more affluent locations, say the waterfront, will, on average, make more revenue. However, when analyzing all the scraped LTR listings, we found that the wards of `r wards_higher_bound`, were the wards with the highest asking rent on average. We also found that the matched listings from these wards were listed at a higher rent, independently of the size and the financial performance of their respective STR listings. So, we added this location variable to our model. 

```{r statistical_model, cache = TRUE, cache.lazy = FALSE, cache.extra = mtime(here("output", "str_processed.qsm"))}

listings_pred <- read_csv("output/listings_pred.csv")

mean_pred_bedrooms <- 
listings_pred %>% 
  filter(`Number of bedrooms` <= 4) %>% 
  left_join(rename(select(ward_price_bounds, ward, geo_quant), Ward = ward), by = "Ward") %>% 
  group_by(`Number of bedrooms`, geo_quant) %>% 
  summarize(`Mean asking rent predicted` = mean(`Predicted asking rent`)) %>% 
  spread(geo_quant, `Mean asking rent predicted`) %>% 
  rename(`Lower-cost ward` = lower,
         `Higher-cost ward` = higher)

rm(listings_pred)

```

```{r model-table, results = "asis", include=T}

qload("output/listings_pred_model.qs")

library(stargazer)
names(model$coefficients) <- c('Constant', '2020 Monthly revenue','Bedrooms','Geography')

stargazer(model, type = "latex", single.row = T, dep.var.labels = "Asking rent")

```

Table \@ref(tab:table) presents the statistical model that was used to predict the rent to which a property owner would accept to move his STR property to the LTR market. The constant is \$`r round(as.numeric(model$coefficients["Constant"]), digit = 2)`, and every dollars earned monthly in 2020 rises the asking price by \$`r round(as.numeric(model$coefficients["2020 Monthly revenue"]), digit = 2)`. Moreover, every bedroom accounts for an augmentation of \$`r round(as.numeric(model$coefficients["Bedrooms"]), digit = 2)` in the predicted asking rent. Finally, if the property is located in a more affluent ward, the ones named above, the asking rent is predicted to be higher by an amount of \$`r round(as.numeric(model$coefficients["Geography"]), digit = 2)`.

The model was applied to the entire population of entire home STR in Toronto and has been given to the City. As an example, Table \@ref(tab:tab-pred) presents the mean asking rent predicted for each size of units and the ward category it is located (lower-cost ward vs higher-cost ward). Our estimations of asking rent for all listings also depends on financial performance of each listing. 

``` {r tab-pred, include = TRUE}

library(kableExtra)

mean_pred_bedrooms %>% 
  mutate(across(where(is.numeric), round, 2)) %>%  
  kbl(caption = "Mean asking rent predicted by number of bedrooms and location", align = "lrrrr") %>% 
  kable_styling()


```

\newpage
